---
title: "ChIPpeakAnno: annotate, visualize, and compare peak data"
author:
- name: Jianhong Ou
  affiliation: Duke Univeristy, Durham, USA
- name: Kai Hu
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Haibo Liu
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Junhui Li
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Jun Yu
  affiliation: Novo Nordisk Dicerna TRU, Boston, USA
- name: Lihua Julie Zhu
  affiliation: UMass Chan Medical School, Worcester, USA
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: ChIPpeakAnno
link-citations: yes
bibliography: bibliography.bib
abstract: |
  A comprehensive guide to utilizing `r Biocpkg("ChIPpeakAnno")` [@zhu2010;
  @zhu2013] for batch annotation and analysis of genomic interval data from
  various experimental techniques, including
  [ChIP-seq](https://en.wikipedia.org/wiki/ChIP_sequencing),
  [CUT&RUN](https://en.wikipedia.org/wiki/CUT%26Tag_sequencing),
  [CUT&Tag](https://en.wikipedia.org/wiki/CUT%26Tag_sequencing),
  [ATAC-seq](https://en.wikipedia.org/wiki/ATAC-seq), CLIP-seq, and other
  genomic interval-based assays. `r Biocpkg("ChIPpeakAnno")` provides
  comprehensive tools for peak annotation to genomic features (genes,
  transcripts, exons, TSS, bidirectional promoters), enrichment analysis (GO
  terms and pathways), sequence extraction for motif discovery, statistical
  testing of peak overlaps, visualization of binding patterns, and comparison
  of peak profiles across multiple samples or experimental conditions.
vignette: |
  %\VignetteIndexEntry{ChIPpeakAnno: annotate, visualize, and compare peak data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      tidy = FALSE, 
                      fig.width = 6, 
                      fig.height = 6,
                      fig.align = "center")
# Handle the biofilecache error
suppressPackageStartupMessages(library(BiocFileCache))
bfc <- BiocFileCache()
res <- bfcquery(bfc, 
                "annotationhub.index.rds", 
                field = "rname", 
                exact = TRUE) 
bfcremove(bfc, rids = res$rid)
```

```{r libraries, include = FALSE}
# Load all required packages
suppressPackageStartupMessages({
  library(ChIPpeakAnno)
  library(EnsDb.Hsapiens.v86)
  library(EnsDb.Hsapiens.v75)
  library(TxDb.Hsapiens.UCSC.hg18.knownGene)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(biomaRt)
  library(AnnotationHub)
  library(org.Hs.eg.db)
  library(reactome.db)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(UpSetR)
  library(GenomeInfoDb)
  library(motifStack)
  library(ensembldb)
  library(reshape2)
  library(knitr)
})
# R.utils is optional, load if available
if (requireNamespace("R.utils", quietly = TRUE)) {
  suppressPackageStartupMessages(library(R.utils))
}
```

# Introduction
A wide range of experimental techniques, including ChIP-ChIP [@blat1999],
ChIP-seq [@barski2007; @johnson2007], CUT&RUN [@skene2017], CUT&Tag
[@kaya2019], ATAC-seq [@buenrostro2013], CLIP-seq [@licatalosi2008], NAD-seq,
and other genomic interval-based assays, are widely used in molecular biology
research. These methods identify genomic regions of interest (peaks) that
represent protein-DNA interactions, chromatin accessibility patterns,
RNA-protein interactions, and other molecular processes. By analyzing these
peaks, researchers can gain insights into gene regulation, chromatin structure,
and the functional roles of genomic elements. The `r Biocpkg("ChIPpeakAnno")`
package offers comprehensive tools for annotating and analyzing peaks generated
from all these experimental approaches.

The analysis workflow involves several key procedures, as outlined below.
Researchers can also integrate other omics data to gain a more comprehensive
understanding of gene regulation mechanisms. The steps where the `r
Biocpkg("ChIPpeakAnno")` package can be utilized are highlighted in pink boxes.

```{r, fig.cap = "A workflow for ChIP-seq data analysis", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "dataFlow.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

Key terms in the analysis workflow:

+ **Peak calling**: The process of identifying genomic regions with
  significant read enrichment (peaks), which represent locations of the protein
  or epigenetic marker of interest
+ **Peak annotation**: The association of peaks with nearby genomic features,
  typically gene transcripts, to understand their functional implications
+ **Enrichment analysis**: The identification of gene ontology (GO) terms or
  biological pathways that are associated with the annotated genes
+ **Motif analysis**: The identification of enriched DNA patterns within peaks
  to understand the binding preferences of the target protein

Multiple algorithms have been developed to identify peaks from experimental
data [@thomas2017]. After peak calling, the results are typically converted
into standard formats such as
[_BED_](https://en.wikipedia.org/wiki/BED_(file_format)) and its variants.
These formats enable easy visualization in genome browsers, such as the UCSC
Genome Browser, as custom tracks. While this allows investigators to examine
peak proximity to genomic features like promoters, enhancers, and other
regulatory elements, manually navigating through genome browsers becomes
challenging when dealing with the large number of peaks typically distributed
across the genome.

The _Bioconductor_ package `r Biocpkg("ChIPpeakAnno")` is a pioneering tool
designed for batch annotation of peaks from various experimental technologies.
A key feature of `r Biocpkg("ChIPpeakAnno")` is its ability to dynamically
retrieve up-to-date annotations through integration with the `r
Biocpkg("biomaRt")` package, ensuring access to the most current annotation
resources. Users can also provide their own annotation data or utilize existing
annotation packages. The package offers functions to extract sequences
surrounding peaks, which are useful for peak validation via PCR and motif
discovery [@Durinck2005]. Additionally, `r Biocpkg("ChIPpeakAnno")` facilitates
the identification of enriched GO terms and pathway annotations, and includes
helper functions for visualizing binding patterns and comparing peak profiles
across multiple samples or experimental conditions.

Since its initial release in 2010, `r Biocpkg("ChIPpeakAnno")` has undergone
continuous enhancement, with active development and regular feature additions
based on user
feedback^[https://github.com/jianhong/ChIPpeakAnno/blob/devel/NEWS].
For usage-related questions, users are encouraged to search for solutions or
post new questions on the [Bioconductor Support
Site](https://support.bioconductor.org/new/post/), which is actively monitored
by experienced users and developers. Feature requests, bug reports, and other
concerns can be submitted as issues on the
[ChIPpeakAnno](https://github.com/jianhong/ChIPpeakAnno) GitHub repository.
User input and contributions are greatly appreciated and carefully considered.

# Quick demo
This section provides a quick example demonstrating how to use `r
Biocpkg("ChIPpeakAnno")` to annotate peaks identified with _MACS_ and
_MACS2_[@zhang2008]. The workflow involves several steps, which are discussed
in detail in the following sections:

+ Convert peaks to _GRanges_ (Section \@ref(readinpeak))
+ Handle replicates (Section \@ref(handlereplicates))
+ Prepare annotation file (Section \@ref(prepanno))
+ Annotate peaks (Section \@ref(annopeaks))
+ Add other feature IDs (Section \@ref(addids))

Sections \@ref(readinpeak) through \@ref(profilecompare) provide detailed use
cases that demonstrate the versatility of `r Biocpkg("ChIPpeakAnno")` across
various scenarios. Sections \@ref(workflow1) and \@ref(workflow2) present
commonly used analytical pipelines with step-by-step instructions for different
experimental settings.

## Convert peaks to _GRanges_ {#readinpeakdemo}
```{r}
macs_peak <- system.file("extdata", "MACS_peaks.xls", 
                         package = "ChIPpeakAnno")
macs_peak_gr <- toGRanges(macs_peak, format = "MACS")
head(macs_peak_gr, n = 2)
```

## Prepare annotations {#prepannos}
This section demonstrates how to prepare annotation data from
[Ensembl](https://useast.ensembl.org/index.html) packages. Ensembl packages
are generally preferred because they provide more comprehensive and superior
annotation quality. For detailed information on annotation preparation, please
refer to Section \@ref(prepanno).

```{r}
# Use Ensembl annotation package:
ensembl.hs86.transcript <- transcripts(EnsDb.Hsapiens.v86)
```

## Annotate peaks
```{r}
# Use Ensembl annotation package:
# Default parameters: output = "nearestLocation",
# PeakLocForDistance = "start", FeatureLocForDistance = "TSS", maxgap = -1L,
# select = "all", ignore.strand = TRUE
macs_peak_ensembl <- annotatePeakInBatch(macs_peak_gr, 
                                         AnnotationData = ensembl.hs86.transcript)
head(macs_peak_ensembl, n = 2)
```

For detailed information about the metadata columns, please refer to Section
\@ref(findnearest). Note that the metadata does not include a "gene symbol"
column by default. This can be added using the `addGeneIDs` function, as
demonstrated below.

## Add gene symbols
To add _gene symbols_, users can choose between two approaches: the organism
annotation package `r Biocpkg("org.Hs.eg.db")`, or the `r Biocpkg("bioRmart")`
package. The biomaRt approach provides the most up-to-date information but may
be slower since it queries the
[BioMart](https://useast.ensembl.org/info/data/biomart/index.html) databases
in real time. After loading the required package, the `addGeneIDs` function can
be used to add _gene symbols_. The following example demonstrates the `r
Biocpkg("bioRmart")` approach.

```{r}
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "hsapiens_gene_ensembl")

# Use Ensembl annotation package:
macs_peak_ensembl <- addGeneIDs(annotatedPeak = macs_peak_ensembl, 
                                mart = mart,
                                feature_id_type = "ensembl_transcript_id",
                                IDs2Add = "hgnc_symbol")
head(macs_peak_ensembl, n = 2)
```

The _hgnc_symbol_ metadata column has now been successfully added to the
resulting _GRanges_ object. It is essential to specify the correct
`feature_id_type` and `IDs2Add` parameters for the function to work correctly.
By default, `feature_id_type = "ensembl_gene_id"`. 

The following sections provide detailed explanations of each topic.

# Convert peaks to _GRanges_ {#readinpeak}
`r Biocpkg("ChIPpeakAnno")` requires peaks or genomic regions to be
represented as _GRanges_ objects. The package provides the `toGRanges`
function to convert commonly used peak formats, including _BED_ (and its
variants), _GFF_, and _CSV_, into _GRanges_ objects. Users simply need to
specify the input file format using the `format` parameter; the function
automatically handles coordinate system conversion and column mapping. The
following example demonstrates how to convert a _GFF_ file into a _GRanges_
object.
```{r}
# Convert GFF to GRanges
macs_peak_gff <- system.file("extdata", "GFF_peaks_hg38.gff", 
                             package = "ChIPpeakAnno")
macs_peak_gr1 <- toGRanges(macs_peak_gff, format = "GFF", header = FALSE)
head(macs_peak_gr1, n = 2)

# Convert BED to GRanges:
macs_peak_bed <- system.file("extdata", "MACS_output_hg38.bed", 
                             package = "ChIPpeakAnno")

macs_peak_gr2 <- toGRanges(macs_peak_bed, format = "BED", header = FALSE)
head(macs_peak_gr2, n = 2)
```

# Quality control of peaks {#qualitycontrol} 
peak width/score distribution, peak enrichment, peak overlap, peak
concordance, peak reproducibility
to be added soon...

# Handle replicates {#handlereplicates}

## Assess the concordance of peak replicates
Concordant peaks are peaks from different biological replicates that show a
high degree of overlap in both genomic coordinates and signal intensity. The
presence of concordant peaks suggests that the observed signal represents a
true positive rather than random noise or technical artifacts. The concordance
of replicates can be evaluated by visualizing peak overlaps.

### Find overlapping peaks
The following examples demonstrate how to identify overlapping peaks using the
sample data provided in the `r Biocpkg("ChIPpeakAnno")` package. By default,
two peaks are considered "overlapping" if one range has its start or end
strictly inside the other (`maxgap = -1L`). Setting `maxgap = 1000` classifies
two peaks as "overlapping" if fewer than 1000 positions separate them.
Specifying `minoverlap = 100` ensures that only peaks with at least 100
overlapping positions are treated as "overlapping". When `0 < minoverlap < 1`,
the function identifies overlaps based on the percentage of the peak interval
that is covered.
```{r}
# For two samples:
data(peaks1)
data(peaks2)
head(peaks1, n = 2)
head(peaks2, n = 2)
ol <- findOverlapsOfPeaks(peaks1, peaks2)
names(ol)

# For three samples:
data(peaks3)
ol2 <- findOverlapsOfPeaks(peaks1, peaks2, peaks3,
                           connectedPeaks = "min")

# Find peaks with at least 90% overlap:
ol3 <- findOverlapsOfPeaks(peaks1, peaks2, minoverlap = 0.9)
```

The function returns an _overlappingPeaks_ object that contains comprehensive
information about peak overlaps, enabling further visualization and
interpretation:

+ _venn_cnt_: A _VennCounts_ object that can be used to create Venn diagrams
+ _peaklist_: A list of _GRanges_ objects containing all merged overlapping
  peaks or unique peaks
+ _uniquePeaks_: A _GRanges_ object containing all unique peaks
+ _mergedPeaks_: A _GRanges_ object containing all merged overlapping peaks
+ _peaksInMergedPeaks_: A _GRanges_ object containing all peaks from each
  sample that are involved in overlapping peaks
+ _overlappingPeaks_: A list of data frames containing annotations for all
  overlapping peaks
+ _all.peaks_: A list of _GRanges_ objects containing the input peaks with
  formatted rownames

To determine the number of peaks unique to each peak set (i.e., peaks that do
not overlap with any peaks in other sets), use the following code:
```{r}
# For peaks1:
length(ol$peaklist[["peaks1"]])

# For peaks2:
length(ol$peaklist[["peaks2"]])
```

To obtain merged overlapping peaks, there are two options:

+ Use `ol$mergedPeaks` to retrieve all merged overlapping peaks from all peak
  sets in `ol`. For multiple peak sets, this returns all peaks that overlap
  with peaks from any other peak sets.
+ Alternatively, use `ol$peaklist[["peaks1///peaks2"]]`, where
  `"peaks1///peaks2"` represents a _GRanges_ object containing all merged
  overlapping peaks from _peaks1_ and _peaks2_. With only two peak sets,
  `ol$mergedPeaks` is equivalent to `ol$peaklist[["peaks1///peaks2"]]`. For
  three or more peak sets, `ol$peaklist[["peaks1///peaks2///peaks3"]]`
  represents merged overlapping peaks from all three sets. In this case,
  `ol$mergedPeaks` is equivalent to the union of
  `ol$peaklist[["peaks1///peaks2"]]`, `ol$peaklist[["peaks2///peaks3"]]`, and
  `ol$peaklist[["peaks1///peaks2///peaks3"]]`.

### Visualize the overlaps using Venn diagram {#venn} or UpSet plot
The output from `findOverlapsOfPeaks` can be directly passed to
`makeVennDiagram` to create a Venn diagram and evaluate the degree of overlap
between peak sets.
The `makeVennDiagram` function also calculates a P-value to assess whether the
observed overlap is statistically significant. 
```{r, results = "hide", fig.cap = "Venn diagram showing the number of overlapping peaks for two samples", fig.small = TRUE}
# For two samples:
venn <- makeVennDiagram(ol, totalTest = 100)

# For three samples:
venn2 <- makeVennDiagram(ol2, totalTest = 100)

# For peaks overlap at least 90%:
venn3 <- makeVennDiagram(ol3, totalTest = 100, plot.type = "Upset")
```

The `totalTest` parameter specifies the total number of potential peaks
considered in the [hypergeometric
test](https://en.wikipedia.org/wiki/Hypergeometric_distribution). This value
should be larger than the largest number of peaks in the replicates. See
Section \@ref(hypertest) for guidance on selecting `totalTest`. Since P-values
are sensitive to the choice of `totalTest`, we recommend using the permutation
test implemented as `peakPermTest` in `r Biocpkg("ChIPpeakAnno")`. For details,
see Section \@ref(permtest).

### Visualize distribution of relative positions of overlapping peaks
As mentioned earlier, two peaks are considered "overlapping" if their
distances are within `maxgap`. To visualize the distribution of relative
positions of _peaks1_ relative to _peaks2_, create a pie chart using the
`ol$overlappingPeaks` object. This object contains detailed information about
the relative positions of peaks for each pair of peak sets. For example,
`ol$overlappingPeaks[["peaks1///peaks2"]]` represents the relative positions
of overlapping peaks between _peaks1_ and _peaks2_.
```{r}
names(ol$overlappingPeaks)
dim(ol$overlappingPeaks[["peaks1///peaks2"]])
ol$overlappingPeaks[["peaks1///peaks2"]][1:2, ]
unique(ol$overlappingPeaks[["peaks1///peaks2"]][["overlapFeature"]])
pie1(table(ol$overlappingPeaks[["peaks1///peaks2"]]$overlapFeature))
```

The _overlapFeature_ column describes the relative positions of peaks between
_peaks1_ and _peaks2_:

+ _upstream_: The peak from _peaks1_ is located upstream of the peak from
  _peaks2_ and is within a distance of `maxgap`
+ _downstream_: The peak from _peaks1_ is located downstream of the peak from
  _peaks2_ and is within a distance of `maxgap`
+ _overlapStart_: The peak from _peaks1_ overlaps with the start of the peak
  from _peaks2_
+ _overlapEnd_: The peak from _peaks1_ overlaps with the end of the peak from
  _peaks2_
+ _inside_: The peak from _peaks1_ is completely contained within the peak
  from _peaks2_
+ _includeFeature_: The peak from _peaks1_ contains the entire range of the
  peak from _peaks2_. Note that the term _Feature_ here refers to the peak in
  _peaks2_ rather than a _genomic feature_ as used in peak annotation context

Using both a Venn diagram and a pie chart enables a more comprehensive
evaluation of peak concordance among biological replicates. 

## Obtain a consensus peak set from replicates
For experiments with two or more biological replicates, there are several
strategies^[https://ro-che.info/articles/2018-07-11-chip-seq-consensus] to
obtain a consensus peak set:

+ **Take the intersection of peaks**: 
  + Most conservative, reduces noise by considering the most robust and
    consistently enriched peaks
  + May result in small number of peaks with narrower widths
  + _Tools_: `ChIPpeakAnno::findOverlapsOfPeaks`, `BEDTools intersect`,
    `DiffBind`
+ **Take the union of all peaks**: 
  + Least conservative, provides a more comprehensive view of the binding
    landscape
  + Useful for identifying condition-specific or rare binding events
  + May lead to large number of peaks
  + _Tools_: `BEDTools merge`, `ChIPpeakAnno::findOverlapsOfPeaks`
+ **Merge the overlapping peaks**: 
  + May lead to peaks with broader widths
  + _Tools_: `ChIPpeakAnno::findOverlapsOfPeaks`, `BEDTools merge`
+ **Use IDR (Irreproducible Discovery Rate) filtering**:
  + Statistically rigorous method that assesses peak reproducibility based on
    signal strength and ranking consistency across replicates
  + Provides a quantitative measure of peak confidence
  + Recommended by ENCODE and widely used in the field
  + _Note_: IDR is designed for pairwise comparisons. The
    `ChIPpeakAnno::IDRfilter` function handles 2 replicates. For 3+ replicates,
    perform pairwise IDR analyses and integrate results (e.g., take intersection
    of reproducible peaks from all pairs)
  + _Tools_: `ChIPpeakAnno::IDRfilter` (see `?IDRfilter` for details), `idr`
    package, IDR command-line tool
+ **Majority voting** (peaks present in N out of M replicates):
  + Flexible approach that allows peaks to be present in a subset of replicates
    (e.g., 2 out of 3, or 3 out of 4)
  + Balances between intersection (too strict) and union (too lenient)
  + _Tools_: `ChIPpeakAnno::findOverlapsOfPeaks` (with custom filtering),
    `BEDTools multiIntersectBed`, `rmspc` (MSPC method)
+ **Use peak coverage voting to determine a consensus set**:
  + Leverages read coverage information across replicates
  + _Tools_: `r Biocpkg("consensusSeekeR")`
+ **Merge reads before peak calling**:
  + Combines sequencing reads from all replicates into a single dataset before
    peak calling
  + May mask variability between replicates
  + Simpler workflow but potentially less informative about reproducibility
  + _Tools_: `samtools merge` (for BAM files), then any peak caller (e.g.,
    `MACS2`, `MACS3`); or `Genrich` peak caller without merging BAM files
+ **Statistical model-based approaches**:
  + Leverage peak coverage and/or read depth information
  + Include specialized peak callers designed for replicated data
  + Incorporate models that account for variability between replicates
  + _Tools_: `PePr`, `Sierra`, `epigraHMM`, `rmspc` (MSPC), `DiffBind`

Users should select a strategy based on their specific research questions
and data quality. For high-quality datasets with good replicate concordance,
IDR filtering is often recommended as it provides a statistically rigorous
assessment of reproducibility.

# Prepare annotation data {#prepanno}
An annotation file contains genomic coordinates and other relevant information
for various genomic features, such as genes, transcripts, promoters,
enhancers, and more. By comparing peaks with these coordinates, researchers
can identify which features are most enriched or associated with the peaks.
This helps understand the functional relevance of peaks and provides insights
into potential regulatory elements or genes involved. **Note**: the annotation
files should be from the primary assembly (no alternative assemblies) of the
reference genome with the SAME VERSION as that used for read alignment and 
peak calling.

## Sources of commonly used annotations
Popular annotation files originate from two main sources and are typically
stored in tab-delimited formats, such as _GTF_ or _BED_.
```{r, echo = FALSE}
Resource <- c("Ensembl", "NCBI RefSeq")
Generated_by <- c("EMBL-EBI", "NCBI")
Annotation_criteria <- c("Comprehensive (most transcripts)", "Conservative (fewer transcripts)")
Gene_id_name <- c("Ensembl gene ID", "NCBI Gene ID, or entrezGene ID, or entrez ID")
URL <- c("https://ftp.Ensembl.org/pub/", "https://ftp.ncbi.nlm.nih.gov/refseq/")
df <- data.frame(Resource, Generated_by, Annotation_criteria, URL)

kable(df, caption = 'Commonly used annotation resources')
```

Additionally, the [UCSC Genome
Browser](https://hgdownload.soe.ucsc.edu/downloads.html) team provides processed
annotations based on the above resources that can be easily visualized in the
browser. But many of these annotation files are malformatted and may not be
suitable for use in the `r Biocpkg("ChIPpeakAnno")` package.

We recommend using Ensembl annotations because they are typically more
comprehensive and consistent.

## Directly obtain annotation data from _TxDb_ and _EnsDb_ packages {#txdbensdb}
In _Bioconductor_, tab-delimited GTF/GFF files from the above sources are
built into _TxDb_ or _EnsDb_ packages. _EnsDb_ is tailored to _Ensembl_
annotations and contains additional information such as _gene_name_, _symbol_,
and _gene_biotype_, the _TxDb_ objects are typically created from RefSeq or
UCSC Genome Browser hosted annotations. Users can directly obtain the _TxDb_
and _EnsDb_ objects from the Bioconductor packages, such as `r
Biocpkg("EnsDb.Hsapiens.v86")` and `r
Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")`. 

A comprehensive list of pre-built _Bioconductor_-maintained _TxDb_ and
_EnsDb_ packages is available, including `r Biocpkg("EnsDb.Hsapiens.v86")`.
This list is regularly updated and can be found
[here](https://www.bioconductor.org/packages/release/data/annotation/). 

## Obtain annotation data from `AnnotationHub`
Users can also retrieve annotations using the `r Biocpkg("AnnotationHub")`
package. By default, `r Biocpkg("AnnotationHub")` uses a snapshot that matches
the version of _Bioconductor_ being used, which may be slightly more
up-to-date compared to pre-built packages. Users also have the option to switch
to an earlier version^[https://bioconductor.org/packages/release/bioc/vignettes/AnnotationHub/inst/doc/AnnotationHub.html#configuring-annotationhub-objects].
The following examples demonstrate how to obtain annotations using `r
Biocpkg("AnnotationHub")`.

```{r}
ah <- AnnotationHub()

# Obtain EnsDb:
EnsDb_Mmusculus_all <- query(ah, pattern = c("Mus musculus", "EnsDb"))
head(EnsDb_Mmusculus_all, n = 2)
EnsDb_Mmusculus <- EnsDb_Mmusculus_all[["AH53222"]]
class(EnsDb_Mmusculus)

# Obtain TxDb:
TxDb_Mmusculus_all <- query(ah, pattern = c("Mus musculus", "TxDb"))
head(TxDb_Mmusculus_all, n = 2)
TxDb_Mmusculus <- TxDb_Mmusculus_all[["AH52264"]]
class(TxDb_Mmusculus)
```

## Obtain annotations from BioMart databases {#usebiomart}
The `r Biocpkg("biomaRt")` package provides a convenient interface to the
[BioMart](https://useast.Ensembl.org/info/data/biomart/index.html) databases
maintained by _Ensembl_. By querying `r Biocpkg("biomaRt")`, users can access
the latest available annotations from _Ensembl_. See [this
vignette](https://bioconductor.org/packages/release/bioc/vignettes/biomaRt/inst/doc/accessing_Ensembl.html)
for details.

`r Biocpkg("ChIPpeakAnno")` provides a helper function called `getAnnotation`
that simplifies the retrieval of desired annotations by leveraging `r
Biocpkg("biomaRt")`. The following example demonstrates its usage:
```{r, eval = FALSE}
listMarts()
head(listDatasets(useMart("ENSEMBL_MART_ENSEMBL")), n = 2)
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "mmusculus_gene_ensembl")
anno_from_biomart <- getAnnotation(mart, 
                                   featureType = "transcript")
head(anno_from_biomart, n = 2)
```

## Build custom _EnsDb_ and _TxDb_ databases {#buildtxdbensdb}
To create the most up-to-date or custom _TxDb_ and _EnsDb_ objects, users can
utilize functions such as `makeTxDbFromUCSC`, `makeTxDbFromEnsembl`, and
`ensDbFromGRanges` from the `r
Biocpkg("txdbmaker")`^[https://bioconductor.org/packages/txdbmaker]
package and the `r
Biocpkg("Ensembldb")`^[https://bioconductor.org/packages/ensembldb/]
package. **Note**: use the annotation files for the primary assembly 
(no alternative assemblies) for building custom _TxDb_ and _EnsDb_ objects.

## Convert annotations into _GRanges_
To annotate peaks, the selected annotation data must first be converted into a
_GRanges_ object. This can be accomplished using the `toGRanges` function,
accessor functions, or the `getAnnotation` function.

### Use `toGRanges` function
`r Biocpkg("ChIPpeakAnno")` provides a helper function called `toGRanges` that
can convert annotations from various formats, including _GFF_, _BED_, _CSV_,
_TxDb_, and _EnsDb_, into _GRanges_ objects. The following example demonstrates
its usage.

```{r}
anno_ensdb_transcript <- toGRanges(EnsDb.Hsapiens.v86, 
                                   feature = "transcript")
head(anno_ensdb_transcript, n = 2)
```

Similarly, setting `feature = "gene"` retrieves gene annotations in _GRanges_
format.

### Use accessor functions
When working with _TxDb_ or _EnsDb_ objects, annotations can be obtained in
_GRanges_ format using various accessor functions such as `genes` and
`transcripts`, which are designed for easy retrieval of the desired
information.
```{r}
anno_ensdb_transcript <- transcripts(EnsDb.Hsapiens.v86)

head(anno_ensdb_transcript, n = 2)
```

### Use `getAnnotation` function
The `getAnnotation` function returns output in _GRanges_ format. See Section
\@ref(usebiomart) for details.

# Visualize peak distributions
Plotting peak distributions serves as a valuable quality control measure,
providing an overview of peak localization across the genome. Unexpected
distributions may indicate potential data issues. Additionally, users can
select appropriate annotation files based on peak distribution patterns. For
example, if peaks are enriched near promoters, users may focus on annotating
them with nearby transcripts.

## Plot peak distributions relative to genomic features
The `binOverFeature` function can be used to plot the distribution of peak
counts relative to a specific genomic feature. The following example shows the
distribution of peaks in the `macs_peak_gr2` dataset relative to transcription
start sites (TSSs).  
```{r, fig.cap = "Peak count distribution around TSSs", warning = FALSE}
annotation_data <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)
binOverFeature(macs_peak_gr2, 
               featureSite = "FeatureStart",
               nbins = 20,
               annotationData = annotation_data,
               xlab = "peak distance from TSS (bp)", 
               ylab = "peak count", 
               main = "Distribution of aggregated peak numbers around TSS")
```

By default, `featureSite = "FeatureStart"`, meaning that distance is calculated
from the peak to the feature start (i.e., TSS for transcripts). The plot above,
showing the distribution of peaks around TSS, exhibits a signal summit around
TSS, which is a characteristic pattern observed in peaks from transcription
factor binding experiments.

Peak distribution over multiple genomic features, including exon, intron,
enhancer, proximal promoter, 5' UTR, and 3' UTR, can be plotted in a single
bar graph using `assignChromosomeRegion`.
```{r, fig.cap = "Bar graph showing peak distributions over different genomic features"}
chromosome_region <- assignChromosomeRegion(macs_peak_gr2,
                                            TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
                                            nucleotideLevel = FALSE,
                                            precedence = c("Promoters",
                                                           "immediateDownstream", 
                                                           "fiveUTRs", 
                                                           "threeUTRs",
                                                           "Exons", 
                                                           "Introns"))

# optional helper function to rotate x-axis labels for barplot(): 
# ref: https://stackoverflow.com/questions/10286473/rotating-x-axis-labels-in-r-for-barplot
rotate_x <- function(data, rot_angle) {
  plt <- barplot(data, xaxt = "n")
  text(plt, par("usr")[3], 
       labels = names(data), 
       srt = rot_angle, adj = c(1.1,1.1), 
       xpd = TRUE, cex = 0.6)
}

rotate_x(chromosome_region[["percentage"]], 45)
```

By default, `nucleotideLevel = FALSE`, meaning that peaks are treated as
ranges when determining overlaps with genomic features. If a peak intersects
with multiple features, the feature assignment is determined by the order
specified in the `precedence` argument. If `precedence` is not set, counts for
each overlapping feature will be incremented. If `nucleotideLevel = TRUE`, the
peak summit (a single position, not suitable for broad peaks) will be used
when determining overlaps.

## Plot peak distributions over different feature levels
In addition to inspecting peak enrichment patterns by plotting distributions
against genomic features, users can plot distributions over different feature
levels, each containing multiple categories, using the
`genomicElementDistribution`
function. 

+ _Transcript Level_: "geneLevel"
    + "Promoter"
    + "Gene body"
    + "Downstream"
    + "Distal Intergenic"
    
+ _Exon Level_: "Exons"
    + "5' UTR"
    + "CDS"
    + "3' URT"
    + "Other exon"
    
+ _Exon/Intron/Intergenic_: "ExonIntron"
    + "Exon"
    + "Intron"
    + "Intergenic"

Note that peaks can be classified into multiple categories from different
levels, which can result in total percentages of annotated features exceeding
100%. At each level, since a peak spans a genomic range, it may overlap with
multiple feature categories. In such cases, by default `nucleotideLevel = FALSE`,
meaning that precedence is determined by the order listed in the `labels`
argument.

The `genomicElementDistribution` function accepts either a single peak object
or a list of peak objects as input. A single peak object produces a pie chart,
while a list of peak objects produces a bar graph.

### Pie graph with one peak set
```{r, fig.cap = "Pie graph showing peak distributions over features at different levels"}
genomicElementDistribution(macs_peak_gr1, 
                           TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene)
```

As shown, a significant number of peaks originate from promoter regions.

### Bar graph with a list of peak sets
```{r, fig.cap = "Bar graph showing peak distributions over features at different levels"}
macs_peaks <- GRangesList(rep1 = macs_peak_gr1,
                          rep2 = macs_peak_gr2)
genomicElementDistribution(macs_peaks, 
                           TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene)
```

The consistent patterns between "rep1" and "rep2" indicate high correlation
between the replicates.

## Plot peak overlaps for multiple features {#upset}
An [UpSet](https://en.wikipedia.org/wiki/UpSet_Plot) plot can be created to
visualize peak overlaps across multiple genomic features.
```{r, fig.cap = "UpSet plot showing the overlappings of peak distributions among different genomic features"}
res <- genomicElementUpSetR(macs_peak_gr1,
                            TxDb.Hsapiens.UCSC.hg38.knownGene)
upset(res[["plotData"]], 
      nsets = length(colnames(res$plotData)), 
      nintersects = NA)
```

An UpSet plot can be considered a "high-dimensional Venn diagram" that enables
visualization of overlaps for multiple sets. For example, in the plot above,
the feature set "gene body" (from the "Transcript Level") and the feature set
"intron" (from the "Exon/Intron/Intergenic") share the highest number of common
peaks. See `?genomicElementUpSetR` for the definition of `distal_promoter`.

# Annotate peaks {#annopeaks}
With annotation data, users can assign peaks identified in their experiments
to nearby features such as genes and transcripts. This process is known as
peak annotation. `r Biocpkg("ChIPpeakAnno")` provides three main approaches for
peak annotation:

1. **`annotatePeakInBatch`**: A highly flexible function that offers various
   `output` options for peak-centric or feature-centric annotation
2. **`annoPeaks`**: A peak-centric annotation function that uses expanded
   binding regions and overlap detection, particularly useful for
   bidirectional promoter detection
3. **`annotateHierarchically`**: A comprehensive annotation system that
   applies multiple annotation strategies in parallel and prioritizes results
   based on factor-specific binding patterns (see Section \@ref(hierarchical))

For example, peaks can be annotated based on their nearest (`output =
"nearestLocation"`) or overlapping (`output = "overlapping"`) features using
the peak-centric method. Alternatively, peaks can be annotated based on their
relative locations to features using the feature-centric method. For example,
if a peak is located upstream of a gene within a certain distance (e.g.,
promoter
region), that gene can be assigned to the peak (`output = "upstream"`). The
`bindingRegion` option provides additional flexibility in specifying relative
locations. Detailed explanations are provided below.

The choice between peak-centric and feature-centric methods depends on the
research question, although most users initially choose the peak-centric
approach as it is easier to interpret. For comprehensive annotation with
factor-specific prioritization, consider using `annotateHierarchically` (Section
\@ref(hierarchical)).

## Peak-centric method {#peakcentric}
The nearest or overlapping features can be assigned to peaks by setting the
`output` option to the following values:

+ _output_: criteria for assigning features
    + "nearestLocation": (default) Outputs the features nearest to the peaks
      based on the absolute value of the difference between the peak location
      and the feature location, calculated as
      `abs(PeakLocForDistance - FeatureLocForDistance)`
    + "overlapping": Outputs features that overlap with the peaks within the
      maximum distance specified by the `maxgap` parameter
    + "both": Outputs both the nearest feature and overlapping features that
      are not the nearest
    + "shortestDistance": Outputs features with the shortest distance to the
      peaks; the "shortest distance" is determined from either end of the
      feature to either end of the peak
    
Other relevant parameters:

+ _PeakLocForDistance_: Location of the peak used for distance calculation
    + "middle": (recommended) Uses the center of the peak
    + "start": (default) Uses the peak edge with the smaller coordinate
    + "end": Uses the peak edge with the larger coordinate
    + "endMinusStart": Uses "end" if the feature is on the plus strand, and
      "start" if on the minus strand
    
+ _FeatureLocForDistance_: Location of the feature used for distance
  calculation
    + "middle": Uses the center of the feature
    + "start": Uses the feature edge with the smaller coordinate
    + "end": Uses the feature edge with the larger coordinate
    + "TSS": (default) Uses "start" if the feature is on the plus strand and
      "end" if on the minus strand
    + "geneEnd": Uses "end" if the feature is on the plus strand and "start" if
      on the minus strand
    
+ _maxgap_: The maximum _gap_ allowed between the peak and feature ranges for
  them to be considered overlapping; defined as the number of bases separating
  the two ranges, where the _gap_ between two adjacent ranges is 0

The following diagram illustrates how to annotate peaks using the peak-centric
method. When `output = "nearestLocation"`, the distance between the peak and
the feature is calculated as `abs(PeakLocForDistance - FeatureLocForDistance)`.
For demonstration purposes, `PeakLocForDistance = "start"` and
`FeatureLocForDistance = "TSS"` are used.
```{r, fig.cap = "How does peak-centric annotation method work", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "annoPeakCentric.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

## Feature-centric method {#featurecentric}
Peaks can also be annotated based on their relative locations to nearby
features. For example, setting `output = "upstream"` annotates peaks to
features that they are located upstream of.

### Relative peak-to-feature location
The following options can be used to specify the desired relative locations of
peaks to features.

+ _output_: Criteria for assigning features
    + "upstream": Outputs the feature if the peak resides upstream of it
    + "upstream&inside": Outputs the feature if the peak resides upstream of
      it or is contained within it
    + "inside": Outputs the feature if the peak is completely contained within
      it
    + "inside&downstream": Outputs the feature if the peak resides within it or
      downstream of it
    + "downstream": Outputs the feature if the peak resides downstream of it
    + "upstreamORdownstream": Outputs the feature if the peak resides upstream
      or downstream of it

Other relevant parameter:

+ _maxgap_: Allowed _gap_ when determining overlap between two ranges; will be
  ignored if `output = "inside"`

The following diagram illustrates how peaks are annotated using the
feature-centric method.

```{r, fig.cap = "How does feature-centric annotation method work", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "annoFeatureCentric1.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

When using the feature-centric method, the `annotatePeakInBatch` function
outputs the feature as long as the peak range overlaps with the target region,
except when `output = "inside"`. In that case, the entire peak range must be
completely contained within the feature range to be output.

### More customization with `bindingRegion`
The `bindingRegion` parameter, which refers to the regions where the target TF
binds, adds further flexibility to the feature-centric method when defining the
target region. For example, to annotate peaks to features where peaks are
located within specific regions relative to them, such as from 2000bp upstream
to 500bp downstream of TSS (where most promoters are found), set `output =
"overlapping"`, `FeatureLocForDistance = "TSS"`, and `bindingRegion = c(-2000,
500)`. Once `bindingRegion` is specified, the `maxgap` parameter will be
ignored. 

The following diagram demonstrates several examples of how to use
`bindingRegion` to define target regions.

```{r, fig.cap = "How to define target region with bindingRegion", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "annoFeatureCentric2.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

Bidirectional promoters are genomic regions located upstream of the TSS of two
adjacent genes that are transcribed in opposite directions [@adachi2002].
These promoters typically regulate the expression of both genes. To annotate
peaks located in such regions, use `output = "overlapping"`,
`FeatureLocForDistance = "TSS"`, and `bindingRegion = c(-5000, 3000)`. By
default, `select = "all"`, meaning that all overlapping features will be
output. To annotate peaks to features with the nearest bidirectional promoters,
use `output = "nearestBiDirectionalPromoters"` along with `bindingRegion`. In
this setting, at most one feature will be reported from each direction. When
using `output = "nearestBiDirectionalPromoters"`, both `maxgap` and
`FeatureLocForDistance` will be ignored.

To illustrate, we use the `myPeakList` provided by `r Biocpkg("ChIPpeakAnno")`.
Since different major genome releases (e.g., _hg19_ vs _hg38_) may have
variations in feature coordinates, it is highly recommended to use an
annotation file that matches the genome version used when generating the peak
file. In the case of `myPeakList`, the peaks were originally called against
_hg18_, so it is necessary to use a matching annotation file created with
_hg18_. Alternatively, the `rtracklayer::liftOver()` function can be used to
convert `myPeakList` to _hg38_ coordinates. For step-by-step instructions, see
"Step3" in Section \@ref(custompool).

## Hierarchical annotation method {#hierarchical}
The `annotateHierarchically` function provides a comprehensive annotation system
that hierarchically prioritizes annotations from multiple perspectives based on
factor-specific binding patterns. The function applies multiple annotation
strategies in parallel (promoter-focused, gene body-focused, intergenic,
bidirectional promoters), then intelligently ranks and selects the most
appropriate annotation for each peak based on the biological factor type. Unlike
single-method annotation, this approach captures annotations from multiple
perspectives and uses hierarchical prioritization to determine the best result.

### Key features
+ **Hierarchical Prioritization**: Intelligently ranks and selects the best
  annotation from multiple strategies based on factor-specific binding patterns
  (TF, histone marks, Pol II, etc.)
+ **Parallel Annotation Strategies**: Applies multiple annotation strategies
  simultaneously (promoter-focused, gene body-focused, intergenic, bidirectional
  promoters) to capture comprehensive annotations
+ **Special Case Handling**: Automatically handles bidirectional promoters and
  broad peaks covering multiple features

### When to use hierarchical annotation
+ When you need comprehensive annotation from multiple perspectives
+ When factor-specific prioritization is important
+ When dealing with complex binding patterns (bidirectional promoters, broad
  domains)

### When to use single-method annotation
+ When you have a specific annotation question (e.g., "find peaks near TSS")
+ When computational speed is critical
+ When you want full control over annotation parameters

### Example: hierarchical annotation for transcription factors
```{r}
data(myPeakList)
anno_data <- transcripts(EnsDb.Hsapiens.v86)
seqlevelsStyle(anno_data) <- seqlevelsStyle(myPeakList)

# Specify factor type explicitly (required for default prioritization)
# anno_tf <- annotateHierarchically(myPeakList[1:100], 
#                                    annoData = anno_data,
#                                    factor_type = "TF")
# head(anno_tf, n = 2)
```

The function requires specifying the `factor_type` parameter, which determines
the prioritization strategy. Available factor types include: "TF", "H3K4me3",
"H3K4me2", "H3K27ac", "H3K4me1", "H3K36me3", "H3K27me3", "PolII", "RBP"
(RNA-binding proteins), "ATAC" (accessibility), "3end" (3' end factors),
"exon" (exon-specific), "intron" (intron-specific), and "architectural"
(architectural features).

### Return all annotations with priority scores
By default, `annotateHierarchically` returns the best annotation per peak.
To see all annotations with their priority scores, set `return_all = TRUE`:
```{r, eval = FALSE}
# anno_all <- annotateHierarchically(myPeakList[1:100], 
#                                    annoData = anno_data,
#                                    factor_type = "TF",
#                                    return_all = TRUE)
```

This is useful for understanding how different annotation strategies were ranked
for each peak.

### Custom prioritization
Users can also provide a custom prioritization function if the default
factor-specific prioritization does not meet their needs:
```{r, eval = FALSE}
# customPrioritize <- function(annotated_peaks, weight_distance = 1000) {
#     # Simple custom prioritization: prioritize by distance only
#     if ("distanceToSite" %in% colnames(mcols(annotated_peaks))) {
#         distance <- abs(mcols(annotated_peaks)$distanceToSite)
#         annotated_peaks$priority_score <- weight_distance * exp(-distance / 500)
#     } else {
#         annotated_peaks$priority_score <- 0
#     }
#     return(annotated_peaks)
# }
# 
# anno_custom <- annotateHierarchically(myPeakList[1:100], 
#                                        annoData = anno_data,
#                                        prioritization_function = customPrioritize,
#                                        weight_distance = 2000)
```

## Example 1: find the nearest features {#findnearest}
Peaks can be annotated by assigning nearby features to them. This can be
achieved by setting `output = "nearestLocation"`. When this option is used,
the results may include "overlapping" features as long as they are the
nearest ones to the peaks.
```{r}
data(myPeakList)
peak_to_anno <- myPeakList[1:100]
anno_data <- transcripts(TxDb.Hsapiens.UCSC.hg18.knownGene)

annotated_peak <- annotatePeakInBatch(peak_to_anno, 
                                      output = "nearestLocation",
                                      PeakLocForDistance = "start",
                                      FeatureLocForDistance = "TSS",
                                      AnnotationData = anno_data)
head(annotated_peak, n = 2)
```

Here is a breakdown of the options:

+ `output = "nearestLocation"`: (default) Outputs the nearest features
  calculated as `abs(PeakLocForDistance - FeatureLocForDistance)`. The output
  can consist of both "strictly nearest features (non-overlapping)" and
  "overlapping features" as long as they are the nearest.
+ `PeakLocForDistance = "start"` (default): Uses the 5' end (relative to plus
  strand) of the peak to calculate the distance to features.
+ `FeatureLocForDistance = "TSS"` (default): Uses the 5' end (relative to plus
  strand) for features on the plus strand and the 3' end (also relative to plus
  strand) for features on the minus strand to calculate the distance to
  features.

The following describes the resulting metadata columns:

+ _peak_: ID of the peak
+ _feature_: ID of the feature, such as Ensembl gene ID
+ _start_position_, _end_position_, _feature_strand_: Feature coordinates
+ _insideFeature_: Relative location of the peak to the feature
    + "upstream": Peak resides upstream of the feature
    + "downstream": Peak resides downstream of the feature
    + "inside": Peak resides inside the feature
    + "overlapStart": Peak overlaps with the start of the feature
    + "overlapEnd": Peak overlaps with the end of the feature
    + "includeFeature": Peak includes the feature entirely
+ _distancetoFeature_: Peak-to-feature distance calculated as
  `abs(PeakLocForDistance - FeatureLocForDistance)`
+ _shortestDistance_: The shortest distance from either end of the peak to
  either end of the feature
+ _fromOverlappingOrNearest_: Relevant only when `output = "both"`;
  "nearestLocation" indicates that the feature is the closest to the peak and
  can overlap with the peak; "Overlapping" means that the feature overlaps with
  the peak but is not the nearest; when `output = "nearestLocation"`, this
  column should consist only of "NearestLocation"

## Example 2: find the nearest and overlapping features {#findboth}
Additionally, `annotatePeakInBatch` can report both the nearest features and
overlapping features by setting `output = "both"` and the `maxgap` parameter.
For example, the following command outputs the nearest features plus all
overlapping features within 100bp.
```{r}
annotated_peak <- annotatePeakInBatch(peak_to_anno, 
                                      AnnotationData = anno_data,
                                      output = "both",
                                      maxgap = 100)
head(annotated_peak, n = 4)
```

Now, the `fromOverlappingOrNearest` column contains both "NearestLocation" and
"Overlapping" categories.

## Visualize peak-to-feature distances
The relative location distribution of peaks to features can be visualized using
the information stored in the _insideFeature_ metadata column.
```{r, fig.cap = "Peak distribution around features"}
#pie1(table(annotated_peak$insideFeature))
donut(table(annotated_peak$insideFeature))
```

## Use custom annotation data
Users can also create and pass user-defined feature coordinates in _GRanges_
format as `annotationData`. For example, if you have a list of transcription
factor binding sites obtained from literature mining and would like to use them
to annotate peaks, you can convert the TF binding site coordinates into a
_GRanges_ object and then pass that object to `annotatePeakInBatch`.
```{r}
TF_binding_sites <- GRanges(seqnames = c("1", "2", "3", "4", "5", "6", "1", 
                                         "2", "3", "4", "5", "6", "6", "6", 
                                         "6", "6", "5"),
                            ranges = IRanges(start = c(967659, 2010898, 2496700, 
                                                       3075866, 3123260, 3857500,
                                                       96765, 201089, 249670, 
                                                       307586, 312326, 385750, 
                                                       1549800, 1554400, 1565000, 
                                                       1569400, 167888600),
                                             end = c(967869, 2011108, 2496920, 
                                                     3076166, 3123470, 3857780,
                                                     96985, 201299, 249890, 307796, 
                                                     312586, 385960, 1550599, 
                                                     1560799, 1565399, 1571199, 
                                                     167888999),
                                             names = paste("t", 1:17, sep = "")),
                            strand = c("*", "*", "*", "*", "*", "*", "*", "*", "*", 
                                       "*", "*", "*", "*", "*", "*", "*", "*"))

annotated_peak2 <- annotatePeakInBatch(peaks1, AnnotationData = TF_binding_sites)
head(annotated_peak2, n = 2)
```

Another example of using user-defined `AnnotationData` is to annotate peaks by
promoters. A promoter is typically defined as the DNA sequence located
immediately upstream of the TSS of a gene. The specific size of a promoter can
vary depending on the gene, its regulatory complexity, and the species being
studied. In practice, the promoter region is often defined as 2000bp upstream
and 500bp downstream from the TSS. To prepare a custom annotation file
containing only promoters, users can use the accessor function `promoters`.
Similar results can be obtained using the feature-centric approach described in
Section \@ref(featurecentric).
```{r}
promoter_regions <- promoters(TxDb.Hsapiens.UCSC.hg18.knownGene, 
                              upstream = 2000, downstream = 500)
head(promoter_regions, n = 2)

annotated_peak3 <- annotatePeakInBatch(peak_to_anno, 
                                       AnnotationData = promoter_regions)
head(annotated_peak3, n = 2)
```

# Add other feature IDs {#addids}
Depending on the annotation file used, the features assigned to peaks may have
different feature IDs. For example, if peaks are annotated with genes using
the `TxDb.Hsapiens.UCSC.hg38.knownGene` annotation file, the feature ID
provided will be "entrez ID". Conversely, if peaks are annotated using the
`EnsDb.Hsapiens.v86` annotation file, the feature ID will be "Ensembl gene ID". 
```{r}
anno_txdb <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
head(anno_txdb$gene_id, n = 5)

anno_ensdb <- genes(EnsDb.Hsapiens.v86)
head(anno_ensdb$gene_id, n = 5)
```

The feature ID from the annotation file will be listed under the "feature"
metadata column in the annotated peak _GRanges_ object. To link these feature
IDs to other IDs such as "symbol", use the `addGeneIDs` function.

The `addGeneIDs` function can accept either a vector of feature IDs or an
annotated peak _GRanges_ object as input. It works by creating a mapping
between the input feature IDs and the IDs to be linked, using either an
organism annotation dataset (`OrgDb` object, such as `org.Hs.eg.db`) or a
_BioMart_ dataset (`Mart` object, such as `useMart(biomart = "Ensembl",
dataset = "hsapiens_gene_Ensembl")`).

To use the function correctly, provide the input feature ID type using the
`feature_id_type` argument, and specify the feature ID types to be linked via
the `IDs2Add` argument. The supported `feature_id_type` and `IDs2Add` options
differ between `OrgDb` and `Mart`. The following summarizes commonly used ID
types.

+ Use _OrgDb_
    + _feature_id_type_: Input feature ID type
        + "ensembl_gene_id"
        + "entrez_id"
        + "gene_symbol"
        + "gene_alias"
        + "refseq_id"
    + _IDs2Add_: Feature ID types to be linked
        + "ensembl"
        + "ensembltrans"
        + "ensemblprot"
        + "entrez_id"
        + "refseq"
        + "symbol"
        
+ Use _Mart_
    + _feature_id_type_: Input feature ID type; use `listFilters(mart)` to see
      the full list
        + "ensembl_gene_id"
        + "ensembl_transcript_id"
        + "ensembl_exon_id"
        + "external_gene_name"
        + "entrezgene_id"
    + _IDs2Add_: Feature ID types to be linked; use `listAttributes(mart)` to
      see the full list
        + "ensembl_gene_id"
        + "ensembl_transcript_id"
        + "ensembl_exon_id"
        + "gene_biotype"
        + "transcript_biotype"
        + "entrezgene_id"
        + "hgnc_symbol"
        + "entrezgene_trans_name"

## Example1: find gene symbols for a vector of entrez IDs
The following example demonstrates how to use the `addGeneIDs` function to
find gene symbols for a vector of entrez IDs using `OrgDb`. Note that the
`"org.Hs.eg.db"` package name must be quoted.
```{r}
ucsc.hg38.knownGene <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
entrez_ids <- head(ucsc.hg38.knownGene$gene_id, n = 10)
print(entrez_ids)

res <- addGeneIDs(entrez_ids, 
                  orgAnn = "org.Hs.eg.db", 
                  feature_id_type = "entrez_id",
                  IDs2Add = "symbol")
head(res, n = 3)
```

## Example2: add gene symbols to annotated peaks
For this example, we annotate `macs_peak_gr2` (obtained in Section
\@ref(readinpeak)) using transcript information from _TxDb_, and then add gene
symbols to the annotated peaks.
```{r}
txdb.hg38.transcript <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)
head(txdb.hg38.transcript, n = 4)
head(names(txdb.hg38.transcript), n = 4)
```

The `txdb.hg38.transcript` object contains a metadata column called _tx_name_,
which holds the Ensembl transcript IDs. However, the `annotatePeakInBatch`
function requires this information to be stored as the names of the
`txdb.hg38.transcript` object (`names(txdb.hg38.transcript)`). This is
necessary to generate annotated peaks that are compatible with the `addGeneIDs`
function. To achieve this, extract the transcript IDs and assign them as names,
as shown below.
```{r}
tr_id <- txdb.hg38.transcript$tx_name
tr_id <- sub("\\..*$", "", tr_id) # get rid of the trailing version number
names(txdb.hg38.transcript) <- tr_id
head(txdb.hg38.transcript, n = 4)
```

Now, we can annotate `macs_peak_gr2` with `annotatePeakInBatch`.
```{r}
res <- annotatePeakInBatch(macs_peak_gr2, 
                           AnnotationData = txdb.hg38.transcript)
head(res, n = 2)
```

As anticipated, the resulting `feature` column contains Ensembl transcript IDs. 
Next, we utilize the `mart` option to add gene symbols.
```{r}
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "hsapiens_gene_ensembl")
res <- addGeneIDs(res, 
                  mart = mart,
                  feature_id_type = "ensembl_transcript_id",
                  IDs2Add = "hgnc_symbol")
head(res, n = 2)
```

Use `listMarts` to show available `biomart` options and `listDatasets` to show
available `dataset` options. Note that, unlike the `OrgDb` option, `hgnc_symbol`
must be supplied instead of `symbol` for the `IDs2Add` argument.

# Enrichment analysis
Enrichment analysis is a crucial step in determining whether specific
biological processes, pathways, or functional categories are over-represented
among the genes associated with peaks. This analysis provides functional
implications for annotated peaks. Two commonly used methods for enrichment
analysis are gene ontology (GO) and pathway enrichment.

GO is a structured vocabulary that categorizes genes and their products into
three main categories: _Molecular Function_ (what it does), _Biological
Process_ (how it does it), and _Cellular Component_ (where it does it). A
pathway, on the other hand, refers to a set of predefined genes involved in a
coordinated sequence of molecular events or cellular processes that collectively
perform a specific biological function. Examples of biological pathways include
the "Glycolysis pathway", which breaks down glucose into pyruvate, and the
"MAPK signaling pathway", which is involved in cell proliferation,
differentiation, and response to external stimuli. While GO enrichment analysis
provides more general insights, pathway analysis focuses specifically on
predefined pathways. Both methods are commonly used and can be performed with
the `getEnrichedGO` and `getEnrichedPATH` functions in `r
Biocpkg("ChIPpeakAnno")`.

In the following demonstration, we first annotate `macs_peak_gr2` (obtained in
Section \@ref(readinpeakdemo)) using genes from _EnsDb_, then perform GO and
pathway enrichment analysis.
```{r, fig.cap = "Histogram showing enriched GO terms"}
anno_data <- genes(EnsDb.Hsapiens.v86)
annotated_peak4 <- annotatePeakInBatch(macs_peak_gr2,
                                       AnnotationData = anno_data,
                                       output = "both")

enriched_go <- getEnrichedGO(annotated_peak4, 
                             orgAnn = "org.Hs.eg.db", 
                             feature_id_type = "ensembl_gene_id")
enrichmentPlot(enriched_go, label_wrap = 60)
```

Ensure that `"org.Hs.eg.db"` is enclosed in quotes and that `feature_id_type`
matches the ID type stored in the `feature` metadata column of the annotated
peak object. If using genes from _TxDb_ for peak annotation, the `feature` is
likely `entrez_id`. If using genes from _EnsDb_ for annotation, the `feature`
should be `ensembl_gene_id`. The code snippet below shows the number of
enriched GO terms in each category: "bp" for "biological process", "cc" for
"cellular component", and "mf" for "molecular function". By default,
`multiAdjMethod = NULL`, meaning that no multiple testing correction is
performed.
```{r}
length(enriched_go[["bp"]][["go.id"]])
length(enriched_go[["cc"]][["go.id"]])
length(enriched_go[["mf"]][["go.id"]])
```

Using the subset of annotated `macs_peak_gr2` results in zero enriched GO terms
in the "bp" and "cc" categories, and 6 hits in the "mf" category. This
suggests that there are no significantly enriched "bp" or "cc" terms associated
with this subset of peaks. However, there are 6 significantly enriched "mf"
terms.
```{r}
head(enriched_go[["mf"]], n = 2)
```

Pathway enrichment analysis can be performed using popular databases such as
_Reactome_ and _KEGG (Kyoto Encyclopedia of Genes and Genomes)_. _Reactome_ is
renowned for its detailed and expert-curated information, while _KEGG_ offers
a broader scope of information, including pathways, diseases, drugs, and more
organisms. The `getEnrichedPATH` function can use either database by specifying
the `pathAnn` parameter. 

For demonstration, we use the built-in dataset `annotatedPeak` with the
_Reactome_ database. Note that the `feature_id_type` for this dataset is
`ensembl_gene_id`. To switch to the _KEGG_ database, simply set `pathAnn =
"KEGGREST"`.
```{r, fig.cap = "Bar graph showing enriched pathways"}
data(annotatedPeak)
enriched_path <- getEnrichedPATH(annotatedPeak,
                                 orgAnn = "org.Hs.eg.db",
                                 feature_id_type = "ensembl_gene_id",
                                 pathAnn = "reactome.db")
```

The enriched pathways can be visualized using the `enrichmentPlot` function.
```{r, fig.cap = "Histogram showing enriched pathways"}
enrichmentPlot(enriched_path)
```

Using `getEnrichedGO` and `getEnrichedPATH` requires an `OrgDb` annotation
package. For less common species that do not have a valid `OrgDb` available,
users can find alternative methods in [this
post](https://support.bioconductor.org/p/9153901/#9153953).

# Motif analysis
Sequence motifs refer to recurring patterns in DNA that are believed to have
biological functions. These motifs often indicate binding sites for proteins
like TFs, while other motifs play roles at the RNA level, such as ribosome
binding and transcription termination. For peaks obtained through experiments
like TF ChIP-seq, motif analysis aids in validating expected binding factors
or discovering transcription factor binding sites (TFBSs), while unanticipated
motifs suggest the presence of co-binding factors.

`r Biocpkg("ChIPpeakAnno")` provides several functions related to motif
analysis, as detailed below:
    + `getAllPeakSequence`: Obtains genomic sequences around peaks
        + Obtained sequences can be used for motif discovery or PCR validation
    + `oligoSummary`: Finds consensus sequences (motifs) in peak sequences
    + `summarizePatternInPeaks`: Checks if given motifs appear in peak
      sequences

## Obtain sequences surrounding the peaks
The following example demonstrates how to retrieve peak sequences, including
20bp upstream and 20bp downstream, for the `macs_peak_gr2` peaks obtained in
Section \@ref(readinpeak).
```{r}
sequence_around_peak <- getAllPeakSequence(macs_peak_gr2, 
                                           upstream = 20,
                                           downstream = 20, 
                                           genome = BSgenome.Hsapiens.UCSC.hg38)
head(sequence_around_peak, n = 2)
```

The `genome` argument can accept either a _BSgenome_ object or a _Mart_ object.
It is important to ensure that the genome version used matches the one used for
creating the peak file. A full list of available _BSgenome_ objects can be found
[here](https://www.bioconductor.org/packages/release/data/annotation/). 

The following example demonstrates how to use the `mart` option. This approach
is slower compared to using a _BSgenome_ because it queries _BioMart_ for
annotations on the fly if `AnnotationData` is not set. See Section
\@ref(usebiomart) for more information about _Mart_.
```{r, eval = FALSE}
mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL",
                dataset = "hsapiens_gene_ensembl")

sequence_around_peak <- getAllPeakSequence(macs_peak_gr2[1], 
                                           upstream = 20,
                                           downstream = 20, 
                                           genome = mart)
```

To save sequences in _FASTA_ format, use the `write2FASTA` function.
```{r, eval = FALSE}
write2FASTA(sequence_around_peak, file = "macs_peak_gr2.fa")
```

## Discover consensus sequences (motifs) in the peaks
The `oligoSummary` function utilizes Markov models to determine if a motif is
enriched in a set of sequences relative to the background. As a prerequisite,
the frequencies of all combinations of short oligonucleotides of a specified
length in the background must first be calculated [@leung1996over]. This can
be accomplished using the `oligoFrequency` function. In the example below, we
aim to identify consensus sequences of length 6.
```{r}
freqs <- oligoFrequency(BSgenome.Hsapiens.UCSC.hg38$chr1)
motif_summary <- oligoSummary(sequence_around_peak, 
                              oligoLength = 6,
                              MarkovOrder = 3,
                              freqs = freqs,
                              quickMotif = TRUE)
```

Here is a breakdown of the arguments:

+ _oligoLength_: the length of the motifs to search for
+ _MarkovOrder_: the order of a Markov chain, which determines how much the 
  current state in the chain depends on previous states; simply put, a 
  zero-order Markov chain does not consider any context or dependencies 
  between adjacent elements, while higher-order Markov chains capture 
  longer-range dependencies in sequences, making them more suitable for 
  modeling complex sequence patterns
+ _freqs_: the background motif frequencies used for the Markov model
+ _quickMotif_: whether or not to generate the motif matrices

The resulting `motif_summary` is a list containing three elements:

+ _zscore_: the Z-score of each oligonucleotide, which quantifies how 
  many standard deviations away the observed count is from the expected
  count; a high positive score suggests that the motif is significantly 
  over-represented in the peaks
+ _counts_: the count numbe of each oligonucleotide
+ _motifs_: a list of motif matrices when `quickMotif = TRUE`

We can use histogram (`hist`) to visualize the resulting Z-scores and 
labels top hits with the `text` function. Below, we label the name of 
the motif that has the highest Z-score.
```{r, fig.cap = "Histogram showing Z-score distribution"}
zscore <- sort(motif_summary$zscore)
h <- hist(zscore, breaks = 100, main = "Histogram of Z-score")
text(x = zscore[length(zscore)], 
     y = h$counts[length(h$counts)] + 1, 
     labels = names(zscore[length(zscore)]), 
     adj = 0, 
     srt = 90)
```

To illustrate how the Z-score can be influenced by the enrichment level of 
various motifs, the following code utilizes simulated data. We first 
generate 5000 sequences with lengths ranging between 100 and 2000 nucleotides. 
Subsequently,we randomly introduce motif 1 into 10% of the sequences and 
motif 2 into 15% of the sequences.
```{r, fig.cap = "Histogram showing Z-score distribution for simulation data"}
set.seed(1)

# motifs to simulate
simulate_motif_1 <- "AATTTA"
simulate_motif_2 <- "TGCATG"

# sample 5000 sequences with lengths ranging from 100 to 2000 nucleotides
# randomly insert motif_1 to 10% of the sequences, and motif_2 to 15% of the sequences
simulation_seqs <- sapply(sample(c(1, 2, 0), 
                                 5000,
                                 prob = c(0.1, 0.15, 0.75),
                                 replace = TRUE), 
                           function(x) {
                             seq <- sample(c("A", "T", "C", "G"),
                                           sample.int(1900, 1) + 100, 
                                           replace = TRUE)
                             insert_pos <- sample.int(length(seq) - 6, 1)
                             if (x == 1) {
                               seq[insert_pos:(insert_pos + 5)] <- strsplit(simulate_motif_1, "")[[1]]
                             } else if (x == 2) {
                               seq[insert_pos:(insert_pos + 5)] <- strsplit(simulate_motif_2, "")[[1]]
                             }
                             paste(seq, collapse = "")
                           }
)

motif_summary_simu <- oligoSummary(simulation_seqs, 
                                   oligoLength = 6, 
                                   MarkovOrder = 3, 
                                   quickMotif = TRUE)
zscore_simu <- sort(motif_summary_simu$zscore, 
                    decreasing = TRUE)
h_simu <- hist(x = zscore_simu, 
               breaks = 100, 
               main = "Histogram of Z-score for simulation data")
text(x = zscore_simu[1:2],  
     y = rep(5, 2), 
     labels = names(zscore_simu[1:2]), 
     adj = 0, 
     srt = 90)
```

As evident from the simulation results, the higher the probability of 
the motif, the larger the resulting Z-score. In addition, you can 
visualize the motif using the `r Biocpkg("motifStack")` package.
```{r, fig.small = TRUE}
pfm <- new("pfm", mat = motif_summary_simu$motifs[[1]],
           name = "sample motif 1")
motifStack(pfm)
```

To loop through each element in `motif_summary$motifs`, we can use the `mapply`
function.
```{r, fig.cap = "Plot showing the first motif detected", fig.small = TRUE}
pfms <- mapply(function(motif, id) { new("pfm", mat = motif, name = as.character(id)) },
               motif_summary$motifs,
               seq_along(motif_summary$motifs))

motifStack(pfms[[1]])
```

## Scan pre-defined sequence patterns in the peaks
If you have a list of motifs (sequence patterns), the `summarizePatternInPeaks`
function can be used to determine whether they are present in the peaks.
```{r}
example_pattern_file <- system.file("extdata/examplePattern.fa",
                                    package = "ChIPpeakAnno")

pattern_in_peak <- summarizePatternInPeaks(
    patternFilePath = example_pattern_file,
    BSgenomeName = BSgenome.Hsapiens.UCSC.hg38,
    peaks = macs_peak_gr2[1:200],
    bgdForPerm = "chromosome",
    nperm = 100, 
    method = "permutation.test")

pattern_in_peak$motif_enrichment
head(pattern_in_peak$motif_occurrence, n = 2)
```

The `summarizePatternInPeaks` function provides two distinct methods (`method =
c("binom.test", "permutation.test")`) for evaluating the significance of given
motif enrichment within target peaks. When utilizing the "binom.test" method,
the expected frequencies for each motif must be determined first, which can be
achieved via `expectFrequencyMethod = c("Markov", "Naive")`. When employing the
"permutation.test" method, users need to provide parameters such as the number
of permutation ("nperm"), the significance level ("alpha"), and methods how
background sequences are determined (`bgdForPerm = c("shuffle", "chromosome"),
chromosome = c("asPeak", "random")`).

The outcome from `summarizePatternInPeaks` comprises two tables. The
"motif_enrichment" table succinctly summarizes the enrichment statistics for
each motif. Specifically, when `method = "binom.test"`, the last column
"pValueBinomTest" represents the P-values resulting from the binomial test.
Conversely, when `method = "permutation.test"`, the last column
"cutOffPermutationTest" denotes the threshold value derived from the
permutation test, determined by the specified significance level. The
"motif_occurrence" table contains detailed information regarding the occurrences
of each motif within every peak.

## Alternative tools
Besides using `r Biocpkg("ChIPpeakAnno")`, users can extract sequences with
the `getAllPeakSequence` function and use other tools such as [MEME
Suite](https://meme-suite.org/meme/doc/streme.html) for motif-related analysis.

# Peak profile comparison {#profilecompare}
When analyzing two or more peak sets from different experiments (e.g., two TFs),
it can be insightful to examine whether peak profiles correlate and how they
compare to each other. `r Biocpkg("ChIPpeakAnno")` offers functions to assess
whether there is significant overlap among multiple peak sets and provides
visualization functions for easy side-by-side comparison of peak patterns.

The significance of overlap can be determined using hypergeometric test or
permutation test, both of which are available in `r Biocpkg("ChIPpeakAnno")`.
However, due to the inherent challenges in estimating the `totalTest` parameter
for the hypergeometric test, we strongly discourage its use.

## Use hypergeometric test to determine overlap among peak sets {#hypertest}
The hypergeometric test is grounded in the principle of the hypergeometric
distribution, which is a probability distribution that describes the likelihood
of drawing a specific number of successes (k) from a sample size of n, given a
finite population (N) that contains K successes when sampling is done without
replacement. The null hypothesis posits that the sample is drawn randomly from
the population, implying that there is no overlap between the two sets of
peaks. A small P-value indicates that the null should be rejected, indicating a
significant overlap between the two sets of peaks.

In `r Biocpkg("ChIPpeakAnno")`, the hypergeometric test is incorporated into
the `makeVennDiagram` function, as detailed in Section \@ref(venn). The
following example illustrates how to compute the hypergeometric test P-values
for peak sets from three TF ChIP-seq experiments.
```{r}
tf1 <- toGRanges(system.file("extdata/TAF.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
tf2 <- toGRanges(system.file("extdata/Tead4.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
tf3 <- toGRanges(system.file("extdata/YY1.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
```

To effectively apply the hypergeometric test, we first need to estimate the
total number of binding sites, i.e. `totalTest`. The selection of `totalTest`
affects the stringency of the test, with smaller values resulting in a more
conservative outcome (larger P-values). For practical guidance on how to choose
an appropriate value for `totalTest`, you can refer to this
[post](https://stat.ethz.ch/pipermail/bioconductor/2010-November/036540.html). 

In our example, we assume that potential binding regions (which include coding
regions and promoter regions) constitute 3% (fairly rough estimation) of the
entire genome. Since the example data is derived from chromosome 2, we can
estimate the number of total binding sites as `(length of chr2) * 3% / (mean
peak length)`.
```{r, results = "hide", fig.cap = "Venn diagram1 showing overlapping peaks", fig.small = TRUE}
overlapping_peaks <- findOverlapsOfPeaks(tf1, 
                                         tf2, 
                                         tf3, 
                                         connectedPeaks = "keepAll")
mean_peak_width <- mean(width(unlist(GRangesList(overlapping_peaks[["all.peaks"]]))))

total_binding_sites <- length(BSgenome.Hsapiens.UCSC.hg38[["chr2"]]) * 0.03 / mean_peak_width
venn1 <- makeVennDiagram(overlapping_peaks, 
                         totalTest = total_binding_sites, 
                         connectedPeaks = "keepAll")
```

For the P-values of each peak pair:
```{r}
venn1[["p.value"]]
```

For the overlapping peak counts:
```{r}
venn1[["vennCounts"]]
```

The parameter `connectedPeaks` denotes how to calculate peak counts when
multiple peaks from different groups overlap. If `connectedPeaks =
"keepFirstListConsistent"`, the counts from the first group will be
consistently displayed in the plot. If `connectedPeaks = "keepAll"`, you will
see multiple numbers: all original peak counts for each group will be displayed
in parentheses, while the count of the minimally involved peaks will be
displayed outside the parentheses. 
```{r, results = "hide", fig.cap = "Venn diagram2 showing overlapping peaks", fig.small = TRUE}
venn2 <- makeVennDiagram(overlapping_peaks, 
                         totalTest = total_binding_sites, 
                         connectedPeaks = "keepFirstListConsistent")
```

Given that all the P-values are extremely small, we must reject the null
hypothesis. This indicates that each pair of peak sets significantly overlaps
with each other. A major drawback of this approach is the necessity to estimate
a `totalTest`, which could dramatically affect the test results. For instance,
if we choose "2%" instead of "3%" in the above example, the P-value for `tf1`
vs. `tf2` increases to 0.49, meaning we can no longer reject the null. To
circumvent the requirement of `totalTest`, we have integrated a permutation
test in the `peakPermTest` function.

## Use permutation test to determine overlap among peak sets {#permtest}
The permutation test is a non-parametric test, which means it does not require
the data to adhere to any specific distribution. The test statistic is
determined according to the observed data, and the null distribution of the
test statistic is estimated using a permutation (re-sampling) procedure. 

In our scenario, the number of overlapping peaks is treated as the test
statistic. Its null distribution is estimated by first re-sampling peaks from a
random peak list (the _peak pool_, which represents all potential binding
sites) followed by counting the number of overlapping peaks. The random peak
list is generated using the distributions found from the input peaks, ensuring
that the peak widths and relative binding positions to the features (such as
_TSS_ and _geneEnd_) follow the same distributions as the input peaks. When the
null hypothesis is valid, the number of overlapping peaks is not significantly
different from what would be expected by chance. Here are some sample codes
using `peakPermTest`.

Example1 demonstrates a permutation test for non-relevant peak sets.
```{r, eval = FALSE}
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
set.seed(123)

# Example1: non-relevant peak sets
utr5 <- unique(unlist(fiveUTRsByTranscript(txdb)))
utr3 <- unique(unlist(threeUTRsByTranscript(txdb)))

utr5 <- utr5[sample.int(length(utr5), 1000)]
utr3 <- utr3[sample.int(length(utr3), 1000)]

pt1 <- peakPermTest(peaks1 = utr3, 
                    peaks2 = utr5,
                    TxDb = txdb, 
                    maxgap = 500,
                    seed = 1)
plot(pt1)
```

```{r, fig.cap = "Permutation test for non-relevant peak sets", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "permTest1.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

Example2 demonstrates a permutation test for highly relevant peak sets.
```{r, eval = FALSE}
# Example2: highly relevant peak sets
cds <- unique(unlist(cdsBy(txdb)))
ol <- findOverlaps(cds, utr3, maxgap = 1)
peaks2 <- c(cds[sample.int(length(cds), 500)],
            cds[queryHits(ol)][sample.int(length(ol), 500)])
pt2 <- peakPermTest(peaks1 = utr3,
                    peaks2 = peaks2,
                    TxDb = txdb,
                    maxgap = 500,
                    seed = 1)
plot(pt2)
```

```{r, fig.cap = "Permutation test for highly relevant peak sets", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "permTest2.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

As observed, for highly relevant peak sets, the P-value is very small. 

### Use custom _peak pool_ {#custompool}
The `peakPermTest` function can automatically generate a _peak pool_ given the
peak binding type (`bindingType = c("TSS", "geneEnd")`), annotation type
(`featureType = c("transcript", "exon")`), and annotation data (_TxDb_).
Additionally, users have the option to construct the _peak pool_ from actual
binding sites derived from experimental data, with hot spots removed. Hot spots
refer to genomic regions that have a high likelihood of being bound by many TFs
in ChIP-seq experiments [@yip2012]. We recommend removing hot spots prior to
the permutation test to prevent overestimation of the association between two
input peak sets. Users are also advised to remove [ENCODE
blacklist](https://github.com/Boyle-Lab/Blacklist/tree/master/lists) regions.
The blacklists were constructed by identifying consistently problematic regions
across independent cell lines and types of experiments for each species in the
_ENCODE_ and _modENCODE_ datasets [@encode2012]. 

Below is an example of creating a _peak pool_ for the human genome using the
TF binding site clusters downloaded from
[ENCODE](https://hgdownload.cse.ucsc.edu/goldenpath/hg38/encRegTfbsClustered/).
The following steps are involved:

+ Step1: download TF binding sites
+ Step2: download hot spots
+ Step3: liftover hot spots to hg38 (if necessary)
+ Step4: select peak sets to test
+ Step5: remove hot spots from binding pool
+ Step6: perform permutation test

```{r, eval = FALSE}
# Step1: download TF binding sites
temp <- tempfile()
download.file(file.path("https://hgdownload.cse.ucsc.edu/",
                        "goldenpath/",
                        "hg38/",
                        "encRegTfbsClustered/",
                        "encRegTfbsClusteredWithCells.hg38.bed.gz"), 
              temp)
df_tfbs <- read.delim(gzfile(temp, "r"), header = FALSE)
unlink(temp)

colnames(df_tfbs)[1:4] <- c("seqnames", "start", "end", "TF")
tfbs_hg38 <- GRanges(as.character(df_tfbs$seqnames),
                     IRanges(df_tfbs$start, df_tfbs$end),
                     TF = df_tfbs$TF)

# Step2: download hot spots
base_url <- "http://metatracks.encodenets.gersteinlab.org/metatracks/"

temp1 <- tempfile()
temp2 <- tempfile()
download.file(file.path(base_url, 
                        "HOT_All_merged.tar.gz"), 
              temp1)
download.file(file.path(base_url, 
                        "HOT_intergenic_All_merged.tar.gz"),
              temp2)
untar(temp1, exdir = dirname(temp1))
untar(temp2, exdir = dirname(temp1))
bedfiles <- dir(dirname(temp1), "bed$")
hot_spots_hg19 <- sapply(file.path(dirname(temp1), bedfiles), toGRanges, format = "BED")
unlink(temp1)
unlink(temp2)

names(hot_spots_hg19) <- gsub("_merged.bed", "", bedfiles)
hot_spots_hg19 <- sapply(hot_spots_hg19, unname)
hot_spots_hg19 <- GRangesList(hot_spots_hg19)

# Step3: liftover hot spots to hg38
temp_chain_gz <- tempfile()
temp_chain <- tempfile()
base_url_chain <- "http://hgdownload.cse.ucsc.edu/goldenpath/"
download.file(file.path(base_url_chain, 
                        "hg19/liftOver/",
                        "hg19ToHg38.over.chain.gz"),
              temp_chain_gz)
gunzip(filename = temp_chain_gz, destname = temp_chain)
chain_file <- import.chain(hg19_to_hg38)
unlink(temp_chain_gz)
unlink(temp_chain)

hot_spots_hg38 <- liftOver(hot_spots_hg19, chain_file)

# Step4: select peak sets to test
tfbs_hg38_by_TF <- split(tfbs_hg38, tfbs_hg38$TF)
TAF1 <- tfbs_hg38_by_TF[["TAF1"]]
TEAD4 <- tfbs_hg38_by_TF[["TEAD4"]]

# Step5: remove hot spots from binding pool
tfbs_hg38 <- subsetByOverlaps(tfbs_hg38, hot_spots_hg38, invert=TRUE)
tfbs_hg38 <- reduce(tfbs_hg38)

# Step6: perform permutation test
pool <- new("permPool", 
            grs = GRangesList(tfbs_hg38), 
            N = length(TAF1))
pt3 <- peakPermTest(TAF1, TEAD4, pool = pool, ntimes = 500)
plot(pt3)
```

```{r, fig.cap = "Permutation test using custom peak pool", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "permTest3.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

## Visualize TSS enrichment signal of multiple experiments {#multiexp}
If you have peak files obtained from multiple TF ChIP-seq experiments and would
like to compare their enriched signals around TSS using raw signals such as
read coverage. `r Biocpkg("ChIPpeakAnno")` provides two functions,
`featureAlignedHeatmap` and `featureAlignedDistribution`, to visualize their
binding patterns side-by-side.

To illustrate this, we first need to prepare both the peak data and coverage
information (_bigWig_). This involves four steps:

+ Step1: read in example peak files
+ Step2: find peaks that are shared by all
+ Step3: read in example coverage files
+ Step4: visualize binding patterns

```{r, fig.cap = "Heatmap of coverage densities for selected TFs"}
# Step1: read in example peak files
extdata_path <- system.file("extdata", package = "ChIPpeakAnno")
broadPeaks <- dir(extdata_path, "broadPeak")
gr_TFs <- sapply(file.path(extdata_path, broadPeaks), toGRanges, format = "broadPeak")
names(gr_TFs) <- gsub(".broadPeak", "", broadPeaks)
names(gr_TFs)
```

Now, we have imported peak files for three TFs ("TAF", "Tead", and "YY1") into
the `gr_TFs` object. The next step is to identify overlapping peaks to ensure
that we are comparing binding patterns over the same genomic regions.
```{r}
# Step2: find peaks that are shared by all
ol <- findOverlapsOfPeaks(gr_TFs)
gr_TFs_ol <- ol$peaklist$`TAF///Tead4///YY1`

# Step3: read in example coverage files
# Here we read in coverage data from -2000bp to -2000bp of each shared peak center.
# Then we extend 2000bp upstream and downstream of the peak center to obtain
# peaks with a uniform length of 4000bp.
gr_TFs_ol_center <- reCenterPeaks(gr_TFs_ol, width = 4000) 

bigWigs <- dir(extdata_path, "bigWig")
coverage_list <- sapply(file.path(extdata_path, bigWigs), 
                        import,  # rtracklayer::import
                        format = "BigWig",
                        which = gr_TFs_ol_center,
                        as = "RleList")

names(coverage_list) <- gsub(".bigWig", "", bigWigs)
names(coverage_list)
```

### Heatmap
```{r, fig.cap = "Heatmap of coverages for selected TFs", fig.height = 6, message = FALSE}
# Step4: visualize binding patterns
sig <- featureAlignedSignal(coverage_list, gr_TFs_ol_center)

# since the bigWig files are only a subset of the original files,
# filter to keep peaks that are with coverage data for all peak sets
keep <- rowSums(sig[[1]]) > 0 & rowSums(sig[[2]]) > 0 & rowSums(sig[[3]]) > 0
sig <- sapply(sig, function(x) x[keep, ], simplify = FALSE)
gr_TFs_ol_center <- gr_TFs_ol_center[keep]

featureAlignedHeatmap(sig, gr_TFs_ol_center,
                      upper.extreme=c(3, 0.5, 4))
```

By default, the rows in the heatmap are ordered by the total coverage per row
from the first sample ("TAF" in this example). We can reorder the rows by
tuning the `sortBy` option. For example, setting `sortBy = "YY1"` will order
the rows by the "YY1" sample in the dataset. You can also sort the rows based
on hierarchical clustering results. Here is a demonstration:
```{r, fig.cap = "Heatmap of coverages for selected TFs sorted by hierarchical clustering", fig.height = 6, message = FALSE}
# perform hierarchical clustering on rows
sig_rowsums <- sapply(sig, rowSums, na.rm = TRUE)
row_distance <- dist(sig_rowsums)
hc <- hclust(row_distance)

# use hierarchical clustering order to sort
gr_TFs_ol_center$sort_by <- hc$order
featureAlignedHeatmap(sig, gr_TFs_ol_center, 
                      upper.extreme = c(3, 0.5, 4),
                      sortBy = "sort_by")
```

### Density plot
Additionally, we can create a density plot using the `featureAlignedDistribution`
function.
```{r, fig.cap = "Distribution of coverage densities for selected TFs", message = FALSE}
featureAlignedDistribution(sig, gr_TFs_ol_center,
                           type = "l")
```

# Common workflow 1: single TF with replicates {#workflow1}
For experiments targeting a single TF with replicates, a common analysis
strategy is outlined below.

+ Step1: import data
  + Convert `BED/GFF` files into _GRanges_
  + Find overlapping peaks among replicates
  + Add metadata (optional)
  + Visualize replicate concordance using a Venn diagram
+ Step2: prepare annotation file
+ Step3: visualize peak distribution
+ Step4: annotate peaks
  + Find possible enhancers
  + Find potential bi-directional promoters
+ Step5: perform enrichment analysis
+ Step6: conduct motif analysis

## Step1: import data
Common peak formats such as _BED_, _GFF_, and _MACS_ can be converted into
_GRanges_ format using the `toGRanges` function. After importing the data,
concordance across peak replicates can be evaluated using `findOverlapsOfPeaks`
and `makeVennDiagram`. Note that metadata columns will be dropped for merged
overlapping peaks. To add them back, use the `addMetadata` function. For
example, `addMetadata(ol, colNames = "score", FUN = mean)` adds a "score"
column to each merged overlapping peak by taking the mean score of individual
peaks involved.
```{r, message = FALSE, warning = FALSE}
# Convert BED/GFF into GRanges
bed1 <- system.file("extdata", "MACS_output_hg38.bed", 
                    package = "ChIPpeakAnno")
gr1 <- toGRanges(bed1, format = "BED", header = FALSE)
gff1 <- system.file("extdata", "GFF_peaks_hg38.gff", 
                    package = "ChIPpeakAnno")
gr2 <- toGRanges(gff1, format = "GFF", header = FALSE)

# Find overlapping peaks
ol <- findOverlapsOfPeaks(gr1, gr2)

# Add "score" metadata column to overlapping peaks
ol <- addMetadata(ol, colNames = "score", FUN = mean) 
head(ol$mergedPeaks, n = 2)
```

```{r, results = "hide", fig.cap = "Venn diagram showing the number of overlapping peaks for gr1 and gr2", fig.small = TRUE}
venn <- makeVennDiagram(ol)
```

For the P-value of hypergeometric test:
```{r}
venn[["p.value"]]
```

For overlapping peak counts:
```{r}
venn[["vennCounts"]]
```

The extremely small P-value suggests high relevance between the two peak sets,
reflecting good consistency among experimental replicates.

## Step2: prepare annotation file
Similar to peak files, the annotation file must also be converted into a
_GRanges_ object. Annotation _GRanges_ can be constructed from _BED_, _GFF_,
and user-defined text files, as well as from _EnsDb_ and _TxDb_ objects using
the `toGRanges` function. For _EnsDb_ and _TxDb_ objects, annotation can also
be prepared using accessor functions, as detailed in Section \@ref(txdbensdb).
Note that the genome version used to create the annotation file must match the
genome used for peak calling, as feature positions may vary across different
genome releases. For example, if using `Mus_musculus.v103` for read mapping,
it is best to use `EnsDb.Mmusculus.v103` for annotation. 

```{r, warning = FALSE}
ensembl.hs86.gene <- toGRanges(EnsDb.Hsapiens.v86, feature = "gene")
head(ensembl.hs86.gene, n = 2)
```

## Step3: visualize peak distribution
### Peak distribution relative to features
Given the merged overlapping peaks and annotation data, the distribution of
distances from the merged overlapping peaks to the nearest features, such as
_genes_ (_TSSs_), can be visualized using the `binOverFeature` function.
```{r, fig.cap = "Peak count distribution around transcription start sites", warning = FALSE}
binOverFeature(ol$mergedPeaks, 
               nbins = 20,
               annotationData = ensembl.hs86.gene,
               xlab = "peak distance from TSSs (bp)", 
               ylab = "peak count", 
               main = "Distribution of aggregated peak numbers around TSS")
```

### Peak distribution over multiple feature levels
The `genomicElementDistribution` function can be used to summarize peak
distribution over different types of genomic features such as exon, intron,
enhancer, and UTR. When inputting a single peak file, a pie graph will be
generated.
```{r, fig.cap = "Pie graph showing peak distributions over different genomic features"}
genomicElementDistribution(ol$mergedPeaks, 
                           TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene)
```

As shown, a significant number of peaks originate from promoter regions, which is
consistent with the signature of peaks obtained from TF binding experiments.
When inputting a list of peak sets (e.g., replicates), a bar graph will be
generated.
```{r, fig.cap = "Bar graph showing peak distributions over different genomic features for two replicates"}
macs_peaks <- GRangesList(rep1 = gr1,
                          rep2 = gr2)
genomicElementDistribution(macs_peaks, 
                           TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene)
```

The bar graph shows that the two peak replicates are consistent. 

### Peak overlappings for multiple features
Peak overlap for multiple feature sets can be visualized using the _UpSet_ plot
(for details, see Section \@ref(upset)).

```{r, fig.cap = "UpSet plot showing peak overlappings for multiple features"}
res <- genomicElementUpSetR(ol$mergedPeak,
                            TxDb.Hsapiens.UCSC.hg38.knownGene)
upset(res[["plotData"]], 
      nsets = length(colnames(res$plotData)), 
      nintersects = NA)
```

## Step4: annotate peaks
Based on the distribution of aggregated peak numbers around the TSS and the
distribution of peaks in different chromosomal regions, most peaks are located
near the TSS. Therefore, it is reasonable to use the feature-centric method to
annotate peaks residing in promoter regions. Promoters can be specified with
the `bindingRegion` parameter. In the following example, the promoter region is
defined as 2000bp upstream and 500bp downstream of the TSS (`bindingRegion =
c(-2000, 500)`).
```{r}
ol_anno <- annotatePeakInBatch(ol$mergedPeak, 
                               AnnotationData = ensembl.hs86.gene, 
                               output = "nearestBiDirectionalPromoters",
                               bindingRegion = c(-2000, 500))
head(ol_anno, n = 2)
```

You can export the annotation to a CSV file:
```{r, eval = FALSE}
ol_anno <- unname(ol_anno)  # remove names to avoid duplicate row.names error
ol_anno$peakNames <- NULL  # remove peakNames to avoid unimplemented type 'list' error
write.csv(as.data.frame(ol_anno), "ol_anno.csv")
```

To visualize the distribution of peaks around features:
```{r, fig.cap = "Pie chart showing peak distribution around features"}
pie1(table(ol_anno$insideFeature))
```

### Find peaks located in bi-directional promoters
In addition to the `output = "nearestBiDirectionalPromoters"` option, `r
Biocpkg("ChIPpeakAnno")` provides a helper function called `peaksNearBDP` to
retrieve statistics for peaks situated in bidirectional promoters. 
```{r}
peaks_near_BDP <- peaksNearBDP(ol$mergedPeaks, 
                               AnnotationData = ensembl.hs86.gene, 
                               MaxDistance = 5000) 
# MaxDistance will be translated into "bindingRegion = 
# c(-MaxDistance, MaxDistance)" internally

peaks_near_BDP$n.peaks
peaks_near_BDP$n.peaksWithBDP
peaks_near_BDP$percentPeaksWithBDP

head(peaks_near_BDP$peaksWithBDP, n = 2)
```

### Find peaks located in enhancers
Enhancers are DNA sequences that can amplify gene expression and are frequently
used as biomarkers for cancer diagnosis and treatment. They can be identified
using a variety of experimental methods such as 3C, 5C, and HiC
[@lieberman2009]. With enhancers obtained through these experimental techniques,
peaks that are located in potential enhancer regions can be identified. The
following example uses 5C data derived from the _hg19_ genome assembly;
therefore, it is necessary to use a matching annotation file.
```{r}
ensembl.hs75.gene <- toGRanges(EnsDb.Hsapiens.v75, feature = "gene")

DNA5C <- system.file("extdata", 
                     "wgEncodeUmassDekker5CGm12878PkV2.bed.gz",
                     package="ChIPpeakAnno")
DNAinteractiveData <- toGRanges(gzfile(DNA5C))

peaks_near_enhancer <-  findEnhancers(peaks = ol$mergedPeaks,
                                      annoData = ensembl.hs75.gene, 
                                      DNAinteractiveData = DNAinteractiveData)
head(peaks_near_enhancer, n = 2)
```

## Step5: perform enrichment analysis
With annotated peaks, the `getEnrichedGO` and `getEnrichedPATH` functions can be
used to retrieve lists of enriched GO or pathway terms, respectively.
```{r, fig.cap = "Histogram showing enriched GO terms"}
enriched_go <- getEnrichedGO(annotatedPeak = ol_anno, 
                             orgAnn = "org.Hs.eg.db", 
                             feature_id_type = "ensembl_gene_id",
                             condense = TRUE)
enrichmentPlot(enriched_go)
```

The following demonstrates pathway enrichment analysis.
```{r, fig.cap = "Histogram showing enriched pathways"}
enriched_pathway <- getEnrichedPATH(annotatedPeak = ol_anno,
                                    orgAnn = "org.Hs.eg.db", 
                                    pathAnn = "KEGGREST",
                                    maxP = 0.05)
enrichmentPlot(enriched_pathway)

# To remove the common suffix " - Homo sapiens (human)":
enrichmentPlot(enriched_pathway, label_substring_to_remove = " - Homo sapiens \\(human\\)")
```

## Step6: conduct motif analysis
To perform motif analysis, sequences surrounding the peaks must first be
extracted, followed by acquiring the consensus sequences. The top motifs
discovered can also be visualized.
```{r}
seq_around_peak <- getAllPeakSequence(ol$mergedPeaks, 
                                      upstream = 20,
                                      downstream = 20, 
                                      genome = BSgenome.Hsapiens.UCSC.hg38)
head(seq_around_peak, n = 2)
```

The `write2FASTA` function can be used to store the result in a FASTA file. The
following code snippet generates Z-scores for short oligos of length 6.
```{r, fig.cap = "Histogram showing Z-score distribution for oligos of length 6"}
freqs <- oligoFrequency(BSgenome.Hsapiens.UCSC.hg38$chr1)
motif_summary <- oligoSummary(seq_around_peak, 
                              oligoLength = 6,
                              MarkovOrder = 3,
                              freqs = freqs,
                              quickMotif = TRUE)
zscore <- sort(motif_summary$zscore)
h <- hist(zscore, 
          breaks = 100, 
          main = "Histogram of Z-score")
text(x = zscore[length(zscore)], 
     y = h$counts[length(h$counts)] + 1, 
     labels = names(zscore[length(zscore)]), 
     adj = 0, 
     srt = 90)
```

The `motifStack` function can be used to visualize the top discovered motifs.
```{r, fig.cap = "Sequence log of detected motif 1", fig.small = TRUE}
pfm <- new("pfm", mat = motif_summary$motifs[[1]],
           name = "sample motif 1")
motifStack(pfm)
```

# Common workflow 2: comparing binding profiles for multiple TFs {#workflow2}
Given two or more peak sets from different TFs, it can be insightful to examine
whether the peak profiles are correlated and how the peak patterns compare to
each other. The workflow presented here demonstrates how to compare the binding
profiles of three TFs. The steps are outlined below.

+ Step1: import data
+ Step2: determine if there is significant overlap among peak sets
+ Step3: visualize and compare the binding patterns
  + Heatmap
  + Density plot

## Step1: import data
```{r, warning = FALSE}
tf1 <- toGRanges(system.file("extdata/TAF.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
tf2 <- toGRanges(system.file("extdata/Tead4.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
tf3 <- toGRanges(system.file("extdata/YY1.broadPeak", package = "ChIPpeakAnno"),
                 format = "broadPeak")
```

## Step2: determine if there is significant overlap among peak sets
To examine associations across different peak sets, `r Biocpkg("ChIPpeakAnno")`
implements both hypergeometric test (`makeVennDiagram`, for details see Section
\@ref(hypertest)) and permutation test (`peakPermTest`, for details see Section
\@ref(permtest)). For demonstration, the hypergeometric test is used here.

```{r, results = "hide", fig.cap = "Venn diagram showing overlapping peaks for three TFs", fig.small = TRUE}
overlapping_peaks <- findOverlapsOfPeaks(tf1, 
                                         tf2, 
                                         tf3, 
                                         connectedPeaks = "keepAll")
mean_peak_width <- mean(width(unlist(GRangesList(overlapping_peaks[["all.peaks"]]))))

total_binding_sites <- length(BSgenome.Hsapiens.UCSC.hg38[["chr2"]]) * 0.03 / mean_peak_width
venn1 <- makeVennDiagram(overlapping_peaks, 
                         totalTest = total_binding_sites, 
                         connectedPeaks = "keepAll")
```

For the P-values of each peak pair:
```{r}
venn1[["p.value"]]
```

Given that the P-values are all extremely low, there is significant overlap
between each pair of peak sets.

## Step3: visualize and compare the binding patterns
Heatmap and density plots can be used to effectively visualize and compare the
binding patterns of multiple TFs. For detailed instructions, see Section
\@ref(multiexp).
```{r, fig.cap = "Heatmap of coverages for selected TFs", fig.height = 6, message = FALSE}
ol_tfs <- findOverlapsOfPeaks(tf1, tf2, tf3, 
                              connectedPeaks = "keepAll")
gr_ol_tfs <- ol_tfs$peaklist$`tf1///tf2///tf3`

TF_width <- width(gr_ol_tfs)
gr_ol_tfs_center <- reCenterPeaks(gr_ol_tfs, width = 4000)

extdata_path <- system.file("extdata", package = "ChIPpeakAnno")
bigWigs <- dir(extdata_path, "bigWig")
coverage_list <- sapply(file.path(extdata_path, bigWigs), 
                        import, # rtracklayer::import
                        format = "BigWig",
                        which = gr_ol_tfs_center,
                        as = "RleList")

names(coverage_list) <- gsub(".bigWig", "", bigWigs)

sig <- featureAlignedSignal(coverage_list, gr_ol_tfs_center)
# since the bigWig files are only a subset of the original files,
# filter to keep peaks that are with coverage data for all peak sets
keep <- rowSums(sig[[1]]) > 0 & rowSums(sig[[2]]) > 0 & rowSums(sig[[3]]) > 0
sig <- sapply(sig, function(x) x[keep, ], simplify = FALSE)
gr_ol_tfs_center <- gr_ol_tfs_center[keep]

featureAlignedHeatmap(sig, gr_ol_tfs_center,
                      upper.extreme=c(3, 0.5, 4))
```

By default, rows are arranged according to the total coverage per row from the
first sample. The following example demonstrates how to sort by the third sample
("YY1") in the dataset by adjusting the `sortBy` option.
```{r, fig.cap = "Heatmap of coverages for selected TFs sorted by YY1 sample", fig.height = 6, message = FALSE}
featureAlignedHeatmap(sig, gr_ol_tfs_center,
                      upper.extreme=c(3, 0.5, 4),
                      sortBy = "YY1")
```

To generate a density plot, use the `featureAlignedDistribution` function.
```{r, fig.cap = "Distribution of coverage densities for selected TFs", message = FALSE}
featureAlignedDistribution(sig, gr_ol_tfs_center, 
                           type="l")
```

As shown, the binding of "YY1" is significantly stronger, while the binding of
"Tead" is considerably weaker.

# Have questions?
For usage-related questions, please post queries on the [Bioconductor Support
Site](https://support.bioconductor.org/new/post/). To report a bug or request a
new feature, please raise an issue on the
[ChIPpeakAnno](https://github.com/jianhong/ChIPpeakAnno) GitHub repository.

# Selected Q & A
## How to import peaks?
`r Biocpkg("ChIPpeakAnno")` provides a helper function, `toGRanges`,
specifically designed for importing files. Users can also use other tools, such
as `rtracklayer::import`.

## How to properly annotate peaks from ChIP-seq data?
`r Biocpkg("ChIPpeakAnno")` offers a variety of options for annotating peaks,
which can be specified using the `output` argument in the `annotatePeakInBatch`
function.

+ To identify the nearest features surrounding peaks, set `output =
  "nearestLocation"`, `output = "shortestDistance"`, or `output =
  "overlapping"`. This is a peak-centric method; for details, see Section
  \@ref(peakcentric).
+ To identify peaks binding to promoter regions, set `output = "upstream"`. For
  more flexibility, adjust the `bindingRegion` option. This is a feature-centric
  method; for details, see Section \@ref(featurecentric).

## Should I use `annotatePeakInBatch` or `annoPeaks`?
Both functions are available and serve different purposes:

+ **`annotatePeakInBatch`**: A comprehensive annotation function that supports
  both point-based and region-based annotation methods. It automatically selects
  the appropriate algorithm based on your parameters:
    + Uses **region-based annotation** (internally calls `annoPeaks`) when
      `bindingRegion` is provided and `bindingType` can be determined
    + Uses **point-based annotation** (internal logic) when `bindingRegion` is
      not provided or cannot be determined
    + Supports multiple `output` options: "nearestLocation", "overlapping",
      "both", "shortestDistance", "upstream", "downstream",
      "nearestBiDirectionalPromoters"
    + Returns all peaks (non-overlapping peaks get NA annotations)

+ **`annoPeaks`**: A specialized region-based annotation function that:
    + Requires `bindingRegion` parameter (e.g., `c(-5000, 5000)`)
    + Uses region expansion and overlap detection
    + Only returns peaks that overlap with expanded annotation regions
      (excludes non-overlapping peaks)
    + Optimized for bidirectional promoter detection
    + Supports `select = "bestOne"` with distance and overlap score
      prioritization

**Recommendation**: For most users, start with `annotatePeakInBatch` as it
provides more flexibility and automatically handles different annotation
scenarios. Use `annoPeaks` directly when you need explicit control over
region-based annotation or when working specifically with bidirectional
promoters. For comprehensive annotation with factor-specific prioritization,
consider using `annotateHierarchically` (Section \@ref(hierarchical)).

## How to prepare annotation data for `annotatePeakInBatch`?
Ensure that the annotation data is in _GRanges_ format for `annotatePeakInBatch`
to work. _TxDb_, _EnsDb_, _GFF_, _BED_, and other formats can be converted into
_GRanges_ using the `toGRanges` function. Additionally, the `getAnnotation`
function can be used to fetch annotation data from _BioMart_. If needed,
personalized annotation files can also be created. For detailed instructions,
see Section \@ref(prepanno).

## Can I output either the nearest or overlapping features?
This question pertains to this [post](https://support.bioconductor.org/p/9154859/#9154869).

When `output = "both"` is set, both "nearest" and "overlapping" features will
be output. If the preference is to assign only one type of feature to each peak
(either "nearest" or "overlapping", with a preference for "overlapping" when
the "nearest" feature is not "overlapping"), the following strategy can be used:
first, annotate peaks with overlapping features; second, annotate peaks lacking
overlapping features with the nearest features; finally, concatenate the two
sets of results. The example code is provided below:
```{r, warning = FALSE}
data(myPeakList)
annoData <- annoGR(EnsDb.Hsapiens.v75)

# Step1: annotate peaks to the overlapping features, if "select = 'all'", multiple
# features can be assigned to a single peak.
anno_overlapping <- annotatePeakInBatch(myPeakList, 
                                        AnnotationData = annoData, 
                                        output = "overlapping", 
                                        select = "first")
anno_overlapping_non_na <- anno_overlapping[!is.na(anno_overlapping$feature)]

# Step2: annotate peaks that are without overlapping features to nearest features
myPeakList_non_overlapping <- myPeakList[!(names(myPeakList) %in% 
                                            anno_overlapping_non_na$peakName)]  
anno_nearest <- annotatePeakInBatch(myPeakList_non_overlapping,
                                    AnnotationData = annoData, 
                                    output = "nearestLocation", 
                                    select = "first")

# Step3: concatenate the two
anno_final <- c(anno_overlapping_non_na, anno_nearest)
```

The provided code assigns either the "overlapping" or "nearest" feature to each
peak. If the "overlapping" feature is not the "nearest", only the "overlapping"
one will be reported.

## Why is the sum of peak numbers in the Venn diagram NOT equal to the sum of the peaks in the original peak lists?
This is a common scenario when calculating the intersection of peaks. Since
peaks represent a range of continuous points rather than single points, they
present a challenge in determining overlapping regions. Consider the
intersection of two lists of range objects: list A (1~3, 4~5, 7~9) and list B
(2~8). The number of overlapping ranges depends on the reference chosen. If list
B is used as the reference, the output would be one range; however, if list A is
used as the reference, the output would be three ranges.

## How does `findOverlapsOfPeaks` count the number of overlapping peaks?
This question is in response to this [post](https://support.bioconductor.org/p/133486/#133603).

When counting the number of overlapping peaks using `findOverlapsOfPeaks`, the
`connectedPeaks` option determines the counting method. If multiple peaks are
involved in any group of connected or overlapping peaks in any input peak list,
setting `connectedPeaks = "merge"` will increment the overlapping counts by one.
Setting `connectedPeaks = "min"` will add the minimal number of involved peaks
in each group of connected or overlapped peaks to the overlapping counts. If
`connectedPeaks = "keepAll"`, it will add the number of involved peaks for each
peak list to the corresponding overlapping counts. Additionally, it will output
counts as if `connectedPeaks` was set to "min". 

For example, if 5 peaks in group1 overlap with 2 peaks in group2:

+ Setting `connectedPeaks = "merge"` will add 1 to the overlapping counts
+ Setting `connectedPeaks = "min"` will add 2 to the overlapping counts
+ Setting `connectedPeaks = "keepAll"` will add 5 peaks to "count.group1", 2 to
  "count.group2", and 2 to the overlapping counts

## Is there a way to show the number of peaks in original peak lists?
Configure `connectedPeaks = "keepAll"` in both the `findOverlapsOfPeaks` and
`makeVennDiagram` functions.

## How to extract the original peak IDs of the overlapping peaks?
In the output of `findOverlapsOfPeaks`, each element in the peak list contains
a metadata column named _peakNames_, which is a CharacterList. This
CharacterList contains the contributing peak IDs with prefixes, e.g.,
"peaks1_peakname1", "peaksi_peaknamej", where "i" denotes peak group i and "j"
denotes the specific peak j within that group. Users can retrieve the original
peak names by splitting these strings. Sample code is provided below:
```{r, warning = FALSE}
# Step1: read in two peak files
bed <- system.file("extdata", 
                   "MACS_output.bed",
                   package = "ChIPpeakAnno")
gff <- system.file("extdata", 
                   "GFF_peaks.gff", 
                   package = "ChIPpeakAnno")
gr1 <- toGRanges(bed, format = "BED", 
                 header = FALSE)
gr2 <- toGRanges(gff, format = "GFF", 
                 header = FALSE, skip = 3)
names(gr2) <- seq_along(gr2) # add names to gr2 for Step4

# Step2: find overlapping peaks
ol <- findOverlapsOfPeaks(gr1, gr2)
peakNames <- ol$peaklist[['gr1///gr2']]$peakNames

# Step3: extract original peak names
peakNames <- melt(peakNames, 
                  value.name = "merged_peak_id") # reshape df
head(peakNames, n = 2)

peakNames <- cbind(peakNames[, 1], 
                   do.call(rbind, 
                           strsplit(as.character(peakNames[, 3]), "__")))

colnames(peakNames) <- c("merged_peak_id", "group", "peakName")
head(peakNames, n = 2)

# Step4: split by peak group
gr1_subset <- gr1[peakNames[peakNames[, "group"] %in% "gr1", "peakName"]]
gr2_subset <- gr2[peakNames[peakNames[, "group"] %in% "gr2", "peakName"]]
head(gr1_subset, n = 2)
head(gr2_subset, n = 2)
```

Alternatively, the following snippet should also work.
```{r}
gr1_renamed <- ol$all.peaks$gr1
gr2_renamed <- ol$all.peaks$gr2
head(gr1_renamed, n = 2)
head(gr2_renamed, n = 2)

peakNames <- melt(ol$peaklist[['gr1///gr2']]$peakNames, 
                  value.name = "merged_peak_id")
gr1_subset <- gr1_renamed[peakNames[grepl("^gr1", peakNames[, 3]), 3]]
gr2_subset <- gr2_renamed[peakNames[grepl("^gr2", peakNames[, 3]), 3]]
head(gr1_subset, n = 2)
head(gr2_subset, n = 2)
```

## How to select the proper number for `totalTest` when using `makeVennDiagram`?
During the evaluation of the association between two sets of peak data using
the hypergeometric distribution, it is essential to determine the total number
of potential binding sites, the `totalTest` parameter in the `makeVennDiagram`.
This parameter should have a value larger than the largest number of peaks in
the peak list. The choice of `totalTest` influences the stringency of the test,
with smaller values yielding more stringent tests and larger P-values. The
computation time for calculating P-values remains independent of the `totalTest`
value. For practical guidance on selecting an appropriate `totalTest` value,
please refer to this
[post](https://stat.ethz.ch/pipermail/bioconductor/2010-November/036540.html).

# How to cite ChIPpeakAnno
If you use `r Biocpkg("ChIPpeakAnno")` in your work, please cite it as follows:
```{r citation, echo = FALSE}
citation(package = "ChIPpeakAnno")
```

# Session info
Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:
```{r sessionInfo, echo = FALSE}
sessionInfo()
```