---
title: "ChIPpeakAnno: A Comprehensive toolkit for Peak Annotation and Analysis"
author:
- name: Jianhong Ou
  affiliation: Duke University, Durham, USA
- name: Haibo Liu
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Kai Hu
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Junhui Li
  affiliation: UMass Chan Medical School, Worcester, USA
- name: Jun Yu
  affiliation: Novo Nordisk Dicerna TRU, Boston, USA
- name: Lihua Julie Zhu
  affiliation: UMass Chan Medical School, Worcester, USA
output:
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 3
  BiocStyle::pdf_document: default
package: ChIPpeakAnno
link-citations: yes
bibliography: bibliography.bib
abstract: |
  This vignette provides a comprehensive workflow guide for using
  `r Biocpkg("ChIPpeakAnno")` to annotate and analyze genomic regions from any
  experimental technique that produces peak-like signals, including DNA-protein
  interactions (ChIP-seq, CUT&RUN, CUT&Tag), RNA-protein interactions
  (RIP-seq, CLIP-seq), RNA-RNA interactions, chromatin accessibility
  (ATAC-seq, DNase-seq), and other genomic interval-based assays. The workflow
  follows a systematic six-phase approach: (1) per-replicate quality assessment
  and filtering, (2) consensus peak creation from multiple replicates, (3)
  factor-specific hierarchical annotation with multiple annotation sources, (4)
  enhancer identification using epigenomic databases, (5) downstream enrichment
  and motif analysis, and (6) visualization and signal analysis. We demonstrate
  the application of key functions including `assessPeaks()`, `filterPeaks()`,
  `findOverlapsOfPeaks()`, `getGenomicAnnotation()`, `annotateHierarchically()`,
  `annotatePeakInBatch()`, `annotatePeaksWithcCRE()`, `annotatePeaksWithRoadmap()`,
  `findEnhancers()`, enrichment analysis functions (`getEnrichedGO()`,
  `getEnrichedPATH()`, `test_enrichment()`, `runLOLA()`), visualization functions
  (`assignChromosomeRegion()`, `metagenePlot()`, `featureAlignedHeatmap()`), and
  integration with sophisticated motif analysis tools (HOMER, STREME, FIMO, memes,
  monaLisa). The vignette also covers annotation data preparation from EnsDb,
  TxDb, biomaRt, and custom sources, as well as best practices for parameter
  selection and common pitfalls.
vignette: |
  %\VignetteIndexEntry{ChIPpeakAnno: A Comprehensive toolkit for Peak Annotation and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  error = FALSE,
  tidy = FALSE, 
  fig.width = 8, 
  fig.height = 6,
  fig.align = "center",
  cache = FALSE
)
```

```{r libraries, include = FALSE}
# Load all required packages
suppressPackageStartupMessages({
  library(ChIPpeakAnno)
  library(GenomicRanges)
  library(IRanges)
  library(GenomeInfoDb)
  library(EnsDb.Hsapiens.v86)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(org.Hs.eg.db)
  library(biomaRt)
})
```

# Introduction
Genomic region analysis has become a cornerstone of modern molecular biology,
encompassing a wide range of experimental techniques that identify and
characterize genomic intervals of interest. These include techniques for
studying DNA-protein interactions (ChIP-ChIP [@blat1999], ChIP-seq
[@barski2007; @johnson2007], CUT&RUN [@skene2017], CUT&Tag), RNA-protein
interactions (RIP-seq, CLIP-seq, PAR-CLIP), RNA-RNA interactions (RIP-seq for
RNA-RNA complexes), chromatin accessibility (ATAC-seq, DNase-seq, FAIRE-seq),
and other genomic assays that produce peak-like signals. Through these
approaches, researchers gain valuable insights into gene regulation, chromatin
structure, RNA processing, and the complex interactions that govern cellular
function. `r Biocpkg("ChIPpeakAnno")` provides a comprehensive toolkit for
annotating and analyzing genomic regions from any of these experimental
approaches, enabling researchers to understand the functional significance of
their identified regions.

## Why Peak Annotation Matters

Peak annotation—the process of associating genomic regions (peaks) with
nearby genomic features, usually gene transcripts, to understand their
functional implications—is a fundamental step in functional genomics
analysis. Proper annotation transforms raw genomic coordinates into
biologically meaningful information, enabling researchers to:

1. **Identify target genes**: Determine which genes are potentially regulated
   by transcription factors, histone modifications, or other chromatin features
2. **Understand regulatory context**: Classify peaks as promoters, enhancers,
   gene bodies, or intergenic regions to understand their functional roles
3. **Enable functional prediction**: Perform gene ontology and pathway
   enrichment analysis to predict biological functions
4. **Integrate with other omics data**: Link binding sites to gene expression
   changes (RNA-seq), epigenetic states, and 3D chromatin structure
5. **Generate testable hypotheses**: Identify candidate target genes for
   experimental validation
6. **Enable comparative analysis**: Compare results across different
   experiments, conditions, or studies using standardized annotation methods

Without proper annotation, peaks remain as abstract genomic coordinates,
making it difficult to interpret their biological significance or integrate
findings with existing knowledge.

## Common Practices for Peak Annotation

The field has established several best practices for peak annotation, largely
based on standards from major consortia (ENCODE, modENCODE) and community
consensus:

### Standard Distance Definitions

+ **Proximal Promoter**: -2000 to +500 bp from transcription start site
  (TSS) - most commonly used standard [@encode2012]
+ **Core Promoter**: -500 to +100 bp from TSS - TATA box region
+ **Extended Promoter**: -5000 to +3000 bp from TSS - for broad factors
+ **Enhancer**: >5 kb from TSS (intergenic or intronic regions)
+ **Intergenic**: >5 kb from any gene

### Annotation Strategies

Three common annotation approaches are supported, each suited to different
analysis goals:

1. **Peak-centric approach**: Find the nearest feature to each peak (e.g.,
   nearest TSS, nearest gene). This approach answers "What is the closest
   gene to each peak?" and is useful for transcription factors with narrow
   binding sites. This is the default approach in many annotation tools
   [@zhu2010].

2. **Feature-centric approach**: Identify peaks within specific genomic
   regions (e.g., promoter regions, gene bodies). This approach answers
   "Which peaks are in promoters?" and is useful for histone marks and broad
   chromatin features. Promoter regions are typically defined as -2 kb to
   +500 bp from TSS [@encode2012].

3. **Hierarchical annotation**: Apply multiple annotation strategies in
   parallel and prioritize results based on factor type. This comprehensive
   approach combines peak-centric and feature-centric methods, then uses
   factor-specific prioritization to select the most biologically relevant
   annotation for each peak. This is implemented in
   \code{annotateHierarchically()} and is recommended for comprehensive
   analysis when factor type is known.

### Quality Considerations

+ **Genome assembly matching**: Always ensure annotation data matches the
  genome assembly used for peak calling (e.g., hg38, mm10)
+ **Use latest annotations whenever possible**: Prefer comprehensive annotation sources
  (GENCODE, Ensembl) and document annotation versions
+ **Factor-specific parameters**: Different factors require different
  annotation strategies with appropriate parameters for each factor type:
  - Transcription factors: Promoter-focused annotation
  - Histone marks: Region-specific or gene body annotation
  - RNA-binding proteins: Strand-aware, transcript-level annotation
+ **Validation**: Always validate annotation results against known
  biological patterns (e.g., promoter enrichment for TFs)

## ChIPpeakAnno
Various algorithms have been developed to identify peaks from experimental
data [@thomas2017]. **Peak calling** identifies genomic regions of
significant read enrichment (peaks), representing binding sites of the
protein or epigenetic marker of interest. While peak files can be
visualized in genome browsers, manual examination of large peak sets is
impractical. The `r Biocpkg("ChIPpeakAnno")` package addresses this need by
providing comprehensive tools for batch annotation and analysis of peaks
from various genomic interval experiments.

`r Biocpkg("ChIPpeakAnno")` has been actively developed since 2010 with
regular feature additions^[https://github.com/jianhong/ChIPpeakAnno/blob/devel/NEWS].
For usage questions, search or post on the [Bioconductor Support
Site](https://support.bioconductor.org/new/post/). For feature requests or
bug reports, open an issue on the
[GitHub repository](https://github.com/jianhong/ChIPpeakAnno).

The analysis involves several key procedures as outlined below. The steps
where `r Biocpkg("ChIPpeakAnno")` package can be utilized are displayed in
pink boxes.
```{r, fig.cap = "A workflow for ChIP-seq data analysis", echo = FALSE, out.width = "100%", out.height = "100%"}
png <- system.file("extdata", "dataFlow.png", package = "ChIPpeakAnno")

knitr::include_graphics(png)
```

**How to cite ChIPpeakAnno**

If you use `r Biocpkg("ChIPpeakAnno")` in your work, please cite it as follows:

```{r citation, echo = FALSE}
citation(package = "ChIPpeakAnno")
```

# Quick demo

This section provides a quick example on utilizing `r Biocpkg("ChIPpeakAnno")`
to annotate peaks identified with _MACS_ or _MACS2_ [@zhang2008]. Detailed
workflow steps are discussed in the following sections.

> **Note**: This quick demo uses example data. For your own analysis, ensure
> your annotation data matches your peak genome assembly (see Section
> \@ref(prepanno) for details).

## Convert peaks to _GRanges_

The `r Biocpkg("ChIPpeakAnno")` package provides an example peak file in
_narrowPeak_ format, which is generated by _MACS_. To work with this example
file, we need to locate it with `system.file` and convert it into a _GRanges_
object using `toGRanges` function.

```{r}
macs_peak <- system.file("extdata", "MACS_peaks.xls", 
                         package = "ChIPpeakAnno")
macs_peak_gr <- toGRanges(macs_peak, format = "MACS")
head(macs_peak_gr, n = 2)
```

## Prepare annotations

This section demonstrates how to prepare annotation file from
[Ensembl](https://useast.ensembl.org/index.html) package. For more detailed
options, see Section \@ref(prepanno).

```{r}
ensembl.hs86.transcript <- transcripts(EnsDb.Hsapiens.v86)
```

## Annotate peaks

```{r}
macs_peak_ensembl <- annotatePeakInBatch(macs_peak_gr,
                                         AnnotationData = ensembl.hs86.transcript,
                                         output = "nearestLocation",
                                         PeakLocForDistance = "middle",
                                         FeatureLocForDistance = "TSS")
head(macs_peak_ensembl, n = 2)
```

# Phase 1: Per-Replicate Processing

> **Overview**: This phase focuses on quality assessment and filtering of
> individual replicates before combining them. Always assess each replicate
> separately to identify potential issues early.

## Step 1: Import Peaks

The first step is to import peak files for each replicate using `toGRanges()`.
This function supports multiple peak calling formats including "BED", "GFF",
"GTF", "MACS", "MACS2", "MACS2.broad", "narrowPeak", "broadPeak", "CSV", and
"others".

> **Tip**: Use `genome(peaks) <- "hg38"` (or your assembly) to set genome
> information immediately after import. This helps with downstream annotation
> and prevents errors.

```{r import-peaks, eval = FALSE}
# Example: Import peaks from different formats
# MACS format
macs_peak <- system.file("extdata", "MACS_peaks.xls", 
                         package = "ChIPpeakAnno")
macs_peaks <- toGRanges(macs_peak, format = "MACS")
```


## Step 2: Quality Assessment (Sample-wise)

Quality assessment is performed **per replicate** using `assessPeaks()`. This
function evaluates multiple quality metrics including peak width distribution,
score metrics (p-value, q-value, signalValue, fold enrichment), and provides
visualizations.

> **Important**: Always assess quality before filtering or annotation. Poor
> quality peaks will lead to poor annotations and misleading biological
> interpretations.

```{r assess-peaks, eval = FALSE}
# Assess each replicate individually
data(Peaks.Ste12.Replicate1)
data(Peaks.Ste12.Replicate2)
data(Peaks.Ste12.Replicate3)

rep1_assessment <- assessPeaks(Peaks.Ste12.Replicate1, 
                               scoreColumns = NULL,
                               verbose = TRUE)

rep2_assessment <- assessPeaks(Peaks.Ste12.Replicate2, 
                               scoreColumns = NULL,
                               verbose = TRUE)

rep3_assessment <- assessPeaks(Peaks.Ste12.Replicate3, 
                               scoreColumns = NULL,
                               verbose = TRUE)
```

```{r assess-peaks-example}
# Assess replicate 1
rep1_assessment <- assessPeaks(Peaks.Ste12.Replicate1, 
                               scoreColumns = NULL, 
                               verbose = FALSE)
cat("Replicate 1 Assessment:\n")
cat("  Median width: ", median(width(Peaks.Ste12.Replicate1)), " bp\n")
cat("  Number of peaks: ", length(Peaks.Ste12.Replicate1), "\n")
```

**Key Quality Metrics:**

- **Peak Width**: 
  - 200-500 bp: Typical for transcription factors
  - 1-5 kb: Typical for histone marks
  - >5 kb: Very broad domains (e.g., H3K27me3)

- **FRiP (Fraction of Reads in Peaks)**: 
  - >0.05 for TFs
  - >0.10 for histone marks

- **NSC/RSC**: 
  - NSC >1.05, RSC >1.0 (ChIPQC standards)

## Step 3: Filter Peaks (Per Replicate)

Based on the assessment results, filter each replicate using `filterPeaks()` to
remove low-quality peaks.

```{r filter-peaks}
# Filter based on width and scores
# Note: Example data may not have score columns, so we filter by width only
peaks1_filtered <- Peaks.Ste12.Replicate1[
  width(Peaks.Ste12.Replicate1) >= 100 & 
  width(Peaks.Ste12.Replicate1) <= 2000]
peaks2_filtered <- Peaks.Ste12.Replicate2[
  width(Peaks.Ste12.Replicate2) >= 100 & 
  width(Peaks.Ste12.Replicate2) <= 2000]
peaks3_filtered <- Peaks.Ste12.Replicate3[
  width(Peaks.Ste12.Replicate3) >= 100 & 
  width(Peaks.Ste12.Replicate3) <= 2000]

cat("After filtering (width 100-2000 bp):\n")
cat("  Replicate 1: ", length(peaks1_filtered), " peaks\n")
cat("  Replicate 2: ", length(peaks2_filtered), " peaks\n")
cat("  Replicate 3: ", length(peaks3_filtered), " peaks\n")

# For data with score columns, use filterPeaks():
# peaks1_filtered <- filterPeaks(Peaks.Ste12.Replicate1, 
#                                 minWidth = 100,
#                                 maxWidth = 2000,
#                                 minScore = 10)
```
# Phase 2: Consensus Peak Creation

> **Overview**: Combine replicates to create robust peak sets. The strategy
> depends on the number of replicates and your research goals (conservative vs
> comprehensive peak sets).

## Step 4: Consensus Peak Creation

For experiments with two or more biological replicates, after filtering each
replicate, users can create consensus peaks per condition. There are several
strategies^[https://ro-che.info/articles/2018-07-11-chip-seq-consensus] to
retrieve a final peak set:

+ Take the intersection of peaks: 
  + Most conservative, reduces noise by considering the most robust and consistently enriched peaks
  + May result in small number of peaks with narrower widths
+ Take the union of all peaks: may lead to large number of peaks
  + Least conservative, provides a broader view of the binding landscape
  + Useful for identifying condition-specific or rare binding events
+ Merge the overlapping peaks: may lead to peaks with broader widths
  + Can be achieved with `ChIPpeakAnno::findOverlapsOfPeaks`
+ Use peak coverage voting to determine a final set:
  + Can be achieved with `r Biocpkg("consensusSeekeR")`
+ Statistical model based approaches that leverage peak coverage and/or read
  depth information

Investigators may choose one of these strategies based on their research
questions and the quality of their data.

For 2 replicates, use `IDRfilter()` (recommended by ENCODE). For 3+ replicates, use
`findOverlapsOfPeaks()` with majority voting.

```{r consensus-peaks}
# For 2 replicates: Use IDR
# Note: IDR requires specific input format
# consensus_peaks <- IDRfilter(peaks1_filtered, peaks2_filtered)

# For 3+ replicates: Use findOverlapsOfPeaks with majority voting
peak_list <- GRangesList(
  rep1 = peaks1_filtered,
  rep2 = peaks2_filtered,
  rep3 = peaks3_filtered
)

# Find peaks present in at least 2 out of 3 replicates
consensus_result <- findOverlapsOfPeaks(peak_list, 
                                       minoverlap = 2,
                                       maxgap = 0L)

consensus_peaks <- consensus_result$mergedPeaks
cat("Consensus peaks (2/3 replicates): ", length(consensus_peaks), "\n")

# View overlap summary
cat("\nOverlap summary:\n")
print(consensus_result$venn_cnt)
```

# Phase 3: Annotation

> **Overview**: This phase assigns peaks to genomic features using
> factor-specific strategies. The choice of annotation method and parameters
> depends on your binding factor type (TF, histone mark, etc.).

## Step 4.5: Prepare Annotation Data {#prepanno}

Before annotating peaks, you need to prepare annotation data that matches your
genome assembly. ChIPpeakAnno supports multiple annotation sources:

### Option 1: Using EnsDb (Recommended)

> **Why EnsDb?**: EnsDb packages provide the most comprehensive annotation
> with gene symbols, biotypes, and metadata. They work seamlessly with
> `annotateHierarchically()`.

EnsDb packages provide annotation with gene symbols and biotypes:

```{r prepare-annotation-ensdb, eval = FALSE}
library(EnsDb.Hsapiens.v86)  # For human hg38
# Or: library(EnsDb.Mmusculus.v79)  # For mouse mm10

# Convert EnsDb to annoGR format
annoData <- annoGR(EnsDb.Hsapiens.v86, feature = "gene")

# For transcript-level annotation
annoData_transcript <- annoGR(EnsDb.Hsapiens.v86, feature = "transcript")
```

**Advantages:**
- Gene and transcript annotations with symbols and biotypes
- Includes metadata
- Works seamlessly with `annotateHierarchically()`

### Option 2: Using TxDb

TxDb packages provide gene model annotations:

```{r prepare-annotation-txdb, eval = FALSE}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Convert TxDb to annoGR format
annoData <- annoGR(TxDb.Hsapiens.UCSC.hg38.knownGene, feature = "gene")

# For transcript-level annotation
annoData_transcript <- annoGR(TxDb.Hsapiens.UCSC.hg38.knownGene, 
                               feature = "transcript")
```

**Note:** TxDb requires an OrganismDb package (e.g., `org.Hs.eg.db`) to get gene
symbols. Use EnsDb if gene symbols are needed.

### Option 3: Using biomaRt (Dynamic Annotation)

biomaRt retrieves up-to-date annotations from Ensembl:

```{r prepare-annotation-biomart, eval = FALSE}
library(biomaRt)

# Connect to Ensembl
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

# Get TSS annotations
TSS <- getAnnotation(mart, featureType = "TSS")

# Get transcript annotations
transcripts <- getAnnotation(mart, featureType = "transcript")

# Get exon annotations
exons <- getAnnotation(mart, featureType = "Exon")
```

**Advantages:**
- Always up-to-date with latest Ensembl releases
- Supports many species
- Can filter by attributes

**Disadvantages:**
- Requires internet connection
- Slower than local annotation packages

### Option 4: Custom GRanges Annotation

You can create custom annotation from your own data:

```{r prepare-annotation-custom, eval = FALSE}
# Create custom annotation from GRanges
custom_annotation <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges = IRanges(start = c(1000, 5000, 2000),
                   end = c(2000, 6000, 3000)),
  strand = c("+", "-", "+"),
  gene_id = c("GENE1", "GENE2", "GENE3"),
  gene_name = c("Gene1", "Gene2", "Gene3")
)

# Convert to annoGR
annoData <- annoGR(custom_annotation)
```

### Choosing the Right Annotation Source

| Source | Best For | Genome Assembly Matching |
|--------|----------|--------------------------|
| EnsDb | Annotation with gene symbols | Must match genome assembly |
| TxDb | Gene models, requires OrganismDb for symbols | Must match genome
assembly |
| biomaRt | Up-to-date annotations, multiple species | Automatically matches |
| Custom | User-defined features | Must match genome assembly |

**Important:** Always ensure your annotation data matches your peak genome
assembly. Use `seqlevelsStyle()` to align chromosome naming if needed.

## Step 5: Preliminary Annotation

Use `getGenomicAnnotation()` to get a quick overview of peak distribution across
genomic features. This helps guide strategy selection for hierarchical
annotation.

```{r preliminary-annotation, eval = FALSE}
# Set up annotation database
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Or use global option
setChIPpeakAnnoGlobals(genome = "hg38", 
                       species = "Homo sapiens",
                       TxDb = txdb)

# Get preliminary annotation
prelim_annotation <- getGenomicAnnotation(consensus_peaks, 
                                          TxDb = txdb,
                                          tssRegion = c(-3000, 3000))

# View distribution
# Access annotated peaks
# View priority annotation distribution
if (!is.null(prelim_annotation$peaks_annotated_with_overlapping_features)) {
  anno_peaks <- prelim_annotation$peaks_annotated_with_overlapping_features
  cat("Annotation distribution:\n")
  print(table(mcols(anno_peaks)$priorityAnnotation, useNA = "ifany"))
  # View the plot
  print(prelim_annotation$plot_priority_annotation)
```

## Step 5.5: Using annotatePeakInBatch (Alternative Annotation Method)

While `annotateHierarchically()` is recommended for most use cases,
`annotatePeakInBatch()` provides flexible annotation with fine-grained control
over annotation parameters. This section explains how to set parameters for
optimal results.

> **When to use `annotatePeakInBatch()`**:
> - You need precise control over distance calculations
> - You want to use specific output modes (e.g., "upstream", "downstream")
> - You're working with custom annotation data
> - You need to combine multiple annotation strategies manually

> **When to use `annotateHierarchically()`**:
> - You want factor-specific intelligent prioritization
> - You want to apply multiple strategies in parallel
> - You prefer a simpler interface with sensible defaults

**For detailed parameter selection guidance**, see the [Interactive Parameter
Guide](docs/Peak_Annotation_Parameter_Guide.md), which provides a
questionnaire-based approach to determine optimal parameters based on your
binding factor type, sequencing method, peak width distribution, and metagene
patterns.

### Key Parameters for annotatePeakInBatch

**1. Output Mode (`output`)**

Controls what type of features are returned:

```{r annotate-peak-in-batch-output, eval = FALSE}
# Nearest location (default) - finds closest feature
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "nearestLocation"  # Default
)

# Overlapping features only
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  maxgap = 0L  # Only true overlaps
)

# Both nearest and overlapping
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "both"
)

# Shortest distance (boundary-to-boundary)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "shortestDistance"
)

# Upstream features (within maxgap of TSS)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "upstream",
  maxgap = 5000  # 5 kb upstream
)

# Bidirectional promoters
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "nearestBiDirectionalPromoters",
  bindingRegion = c(-5000, 3000)
)
```

**2. Reference Points (`PeakLocForDistance` and `FeatureLocForDistance`)**

Controls which part of peaks and features are used for distance calculations:

```{r annotate-peak-in-batch-reference, eval = FALSE}
# For narrow peaks (<200bp): use "start" (default)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  PeakLocForDistance = "start",        # Peak reference point
  FeatureLocForDistance = "TSS"        # Feature reference point
)

# For broad peaks (>200bp): use "middle" (recommended)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  PeakLocForDistance = "middle",       # Use peak center
  FeatureLocForDistance = "TSS"        # Use TSS
)

# For strand-specific assays (e.g., GRO-seq)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  PeakLocForDistance = "endMinusStart", # Strand-aware
  FeatureLocForDistance = "TSS"
)
```

**3. Maximum Gap (`maxgap`)**

Controls how far features can be from peaks:

```{r annotate-peak-in-batch-maxgap, eval = FALSE}
# Strict: only true overlaps (default for overlapping)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  maxgap = -1L  # No gap allowed
)

# Allow adjacent features
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  maxgap = 0L  # Adjacent allowed
)

# Allow features within 5kb
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "upstream",
  maxgap = 5000L  # 5 kb
)
```

**4. Feature Selection (`select`)**

Controls which features to return when multiple matches exist:

```{r annotate-peak-in-batch-select, eval = FALSE}
# Return all matching features
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  select = "all"  # All matches
)

# Return first match only
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  select = "first"  # First match
)

# Return best match (based on distance)
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",
  select = "bestOne"  # Best match
)
```

### Parameter Selection Guide

**For Transcription Factors (narrow peaks, promoter-focused):**

```{r annotate-peak-in-batch-tf, eval = FALSE}
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "nearestLocation",
  PeakLocForDistance = "middle",      # Use peak center
  FeatureLocForDistance = "TSS",     # Distance to TSS
  maxgap = 5000L,                     # Within 5kb
  select = "all"
)
```

**For Histone Marks (broad peaks, gene body):**

```{r annotate-peak-in-batch-histone, eval = FALSE}
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "overlapping",             # Find overlapping features
  PeakLocForDistance = "middle",      # Use peak center
  maxgap = 0L,                        # True overlaps only
  select = "all"                      # All overlapping genes
)
```

**For Enhancer Analysis (upstream regions):**

```{r annotate-peak-in-batch-enhancer, eval = FALSE}
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "upstream",                # Upstream of TSS
  PeakLocForDistance = "middle",
  FeatureLocForDistance = "TSS",
  maxgap = 10000L,                    # Within 10kb upstream
  select = "all"
)
```

**For Bidirectional Promoters:**

```{r annotate-peak-in-batch-bdp, eval = FALSE}
annotated <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "nearestBiDirectionalPromoters",
  bindingRegion = c(-5000, 3000),     # Promoter region
  PeakLocForDistance = "middle"
)
```

### Complete Example

```{r annotate-peak-in-batch-complete, eval = FALSE}
# 1. Prepare annotation data
library(EnsDb.Hsapiens.v86)
annoData <- annoGR(EnsDb.Hsapiens.v86, feature = "gene")

# 2. Annotate with appropriate parameters
annotated_peaks <- annotatePeakInBatch(
  peaks = consensus_peaks,
  AnnotationData = annoData,
  output = "nearestLocation",
  PeakLocForDistance = "middle",      # For broad peaks
  FeatureLocForDistance = "TSS",
  maxgap = 5000L,
  select = "all",
  ignore.strand = TRUE
)

# 3. View results
head(mcols(annotated_peaks))
```

## Step 6: Hierarchical Annotation

Use `annotateHierarchically()` for factor-specific annotation with multiple
parallel strategies. This function applies factor-specific prioritization to
select the best annotation per peak.

```{r hierarchical-annotation, eval = FALSE}
# Set up EnsDb for hierarchical annotation
library(EnsDb.Hsapiens.v86)
ensdb <- EnsDb.Hsapiens.v86

# Or use global option
setChIPpeakAnnoGlobals(EnsDb = ensdb)

# For transcription factors
tf_annotated <- annotateHierarchically(
  peaks = consensus_peaks,
  factor_type = "TF",
  EnsDb = ensdb,
  PeakLocForDistance = "middle",
  bindingRegion = c(-3000, 3000)
)

# For H3K4me3 (promoter mark)
h3k4me3_annotated <- annotateHierarchically(
  peaks = consensus_peaks,
  factor_type = "H3K4me3",
  EnsDb = ensdb,
  PeakLocForDistance = "middle",
  bindingRegion = c(-2000, 500)
)

# For H3K27ac (active enhancer mark)
h3k27ac_annotated <- annotateHierarchically(
  peaks = consensus_peaks,
  factor_type = "H3K27ac",
  EnsDb = ensdb,
  PeakLocForDistance = "middle",
  bindingRegion = c(-5000, 3000)
)
```

```{r hierarchical-annotation-example}
# Set up EnsDb
library(EnsDb.Hsapiens.v86)
ensdb <- EnsDb.Hsapiens.v86
setChIPpeakAnnoGlobals(EnsDb = ensdb)

# Hierarchical annotation for TF
tf_annotated <- annotateHierarchically(
  peaks = consensus_peaks,
  factor_type = "TF",
  PeakLocForDistance = "middle",
  bindingRegion = c(-3000, 3000)
)

cat("TF annotation complete:\n")
cat("  Annotated peaks: ", length(tf_annotated), "\n")
if (length(tf_annotated) > 0 && "gene_name" %in% colnames(mcols(tf_annotated))) {
  cat("  Unique genes: ", length(unique(mcols(tf_annotated)$gene_name[!is.na(mcols(tf_annotated)$gene_name)])), "\n")
```

**Factor-Specific Strategies:**

| Factor Type | Peak Width | Key Parameters |
|------------|------------|----------------|
| TF | 200-500 bp | `factor_type="TF"`, `PeakLocForDistance="middle"` |
| H3K4me3 | 200-500 bp | `bindingRegion=c(-2000, 500)` |
| H3K27ac | 500-2000 bp | `bindingRegion=c(-5000, 3000)` |
| H3K36me3 | 1-5 kb | `bindingType="fullRange"` |
| H3K27me3 | 5-20 kb | `bindingType="fullRange"`, `PeakLocForDistance="middle"`
|

## Step 6.5: Bidirectional Promoter Detection

Bidirectional promoters are shared promoter regions between two head-to-head
(divergently transcribed) genes. Use `peaksNearBDP()` to identify peaks near
these unique regulatory elements:

```{r bidirectional-promoters, eval = FALSE}
# Identify peaks near bidirectional promoters
bdp_peaks <- peaksNearBDP(
  myPeakList = consensus_peaks,
  AnnotationData = annoData,
  maxTSSDistance = 1000L,  # TSSs within 1kb
  PeakLocForDistance = "middle",
  FeatureLocForDistance = "TSS"
)

# View results
head(bdp_peaks, n = 3)
```

**Key Parameters:**
- `maxTSSDistance`: Maximum distance between two divergent TSSs (default: 1000bp)
- `PeakLocForDistance`: Which part of peak to use ("start", "middle", "end")
- `FeatureLocForDistance`: Reference point on features ("TSS", "middle", "geneEnd")

## Step 6.6: Alternative Annotation with annoPeaks

`annoPeaks()` provides an alternative annotation approach using region expansion
and overlap detection, rather than point-to-point distance calculations. This is
useful for queries like "find all peaks within 5kb of any TSS":

```{r anno-peaks, eval = FALSE}
# Annotate using region expansion approach
anno_peaks <- annoPeaks(
  peaks = consensus_peaks,
  annoData = annoData,
  bindingType = "nearestBiDirectionalPromoters",
  bindingRegion = c(-5000, 3000),  # 5kb upstream, 3kb downstream
  select = "bestOne"
)

# Other binding types:
# - "startSite": TSS region
# - "endSite": Gene end region
# - "fullRange": Full gene body
```

**Differences from `annotatePeakInBatch`:**
- Uses region expansion and overlap detection
- Requires `bindingRegion` parameter
- Excludes non-overlapping peaks (doesn't return NA)
- Optimized for bidirectional promoter detection

## Step 6.7: Adding Gene Identifiers

After annotation, use `addGeneIDs()` to enrich peaks with additional gene
identifiers (symbols, Entrez IDs, Ensembl IDs, RefSeq IDs):

```{r add-gene-ids, eval = FALSE}
# Add gene symbols and Entrez IDs
annotated_with_ids <- addGeneIDs(
  annotatedPeak = tf_annotated,
  orgAnn = "org.Hs.eg.db",
  IDs2Add = c("symbol", "entrez_id"),
  feature_id_type = "ensembl_gene_id"
)

# View added columns
head(mcols(annotated_with_ids))
```

**Supported Identifiers:**
- `symbol`: Gene symbols
- `entrez_id`: Entrez gene IDs
- `ensembl_gene_id`: Ensembl gene IDs
- `refseq_id`: RefSeq IDs

# Phase 4: Enhancer Identification

> **Overview**: Link peaks to enhancers and their target genes. Enhancers are
> distal regulatory elements that can be far from their target genes, requiring
> special methods for identification.

## Step 7: Enhancer Identification

Use `findEnhancers()` to identify enhancer-associated peaks. This can use Hi-C
data for 3D chromatin structure or ENCODE CRE annotations.

> **Note**: Enhancer identification requires additional data (Hi-C or CRE
> annotations). If this data is not available, you can still identify
> enhancer-like peaks by annotating with Roadmap Epigenomics or ENCODE cCRE
> data.

```{r enhancer-identification, eval = FALSE}
# Option 1: With Hi-C data (preferred)
if (!is.null(hic_data)) {
  enhancer_results <- findEnhancers(
    peaks = annotated_peaks,
    hic_data = hic_data,
    max_distance = 1e6  # 1 Mb
  )
}

# Option 2: With ENCODE CRE annotations (human/mouse)
# Use annotatePeaksWithcCRE() for enhancer annotation
enhancer_annotated <- annotatePeaksWithcCRE(
  peaks = annotated_peaks,
  species = "Homo sapiens",
  tissue_types = "any"  # or specific tissue types
)
```

```{r enhancer-identification-example}
# Example: Annotate with cCRE data (includes enhancers)
# Note: This requires downloading cCRE data, so we'll show the code structure
cat("Enhancer identification options:\n")
cat("  1. Use findEnhancers() with Hi-C data\n")
cat("  2. Use annotatePeaksWithcCRE() for ENCODE CRE annotations\n")
cat("  3. Use annotatePeaksWithRoadmap() for Roadmap Epigenomics enhancer data\n")
```

## Step 7.5: Listing Available Epigenomic Data

Before downloading epigenomic data, you can list available datasets:

```{r list-epigenomic-data, eval = FALSE}
# List available cCRE datasets
available_ccres <- listAvailablecCREs(species = "Homo sapiens")
head(available_ccres)

# List available Roadmap epigenomes
roadmap_epigenomes <- listAvailableRoadmapEpigenomes(
  data_type = "enhancer"
)
head(roadmap_epigenomes)

# List available Roadmap tissue types
roadmap_tissues <- listAvailableRoadmapTissues()
head(roadmap_tissues)
```

# Phase 5: Downstream Analysis

> **Overview**: After annotation, perform enrichment analysis, gene set
> enrichment, and motif discovery to understand the biological significance of
> your peaks.

## Step 8: Differential Analysis (Multi-Condition Only)

For multi-condition experiments, perform differential peak analysis before
annotation.

> **Important**: Differential analysis should be performed on peak counts
> (using tools like DiffBind or DESeq2) before annotation. Only annotate the
> significant differential peaks to focus on condition-specific changes.

```{r differential-analysis, eval = FALSE}
# Using diffBind (recommended for ChIP-seq)
library(DiffBind)

# Create DBA object
dba <- dba(sampleSheet = "sample_sheet.csv")

# Count reads
dba <- dba.count(dba)

# Normalize
dba <- dba.normalize(dba, method = DBA_DESEQ2)

# Differential analysis
dba <- dba.analyze(dba, method = DBA_DESEQ2)

# Extract differential peaks
diff_peaks <- dba.report(dba, th = 1)

# Filter for significant peaks
sig_diff_peaks <- diff_peaks[abs(diff_peaks$Fold) > 1 & 
                              diff_peaks$FDR < 0.05]

# Then annotate differential peaks using ChIPpeakAnno
annotated_diff <- annotateHierarchically(
  peaks = sig_diff_peaks,
  factor_type = "TF"
)
```

## Step 10: Peak Overlap Enrichment Analysis

**Enrichment analysis** identifies the gene ontology (GO) terms or
biological pathways that are associated with the annotated genes. This
analysis helps understand the functional significance of the peaks by
revealing overrepresented biological processes, molecular functions, and
cellular components.

Compare your peaks against reference peak sets (e.g., ENCODE, Roadmap
Epigenomics) using LOLA or ChIPpeakAnno's overlap functions.

### Visualizing Peak Overlaps with Venn Diagrams

Use `makeVennDiagram()` to visualize overlaps between 2-5 peak sets with
statistical significance testing:

```{r venn-diagram, eval = FALSE}
# Create Venn diagram for 2 peak sets
venn_result <- makeVennDiagram(
  Peaks = list(rep1 = peaks1_filtered, 
               rep2 = peaks2_filtered),
  NameOfPeaks = c("Replicate 1", "Replicate 2"),
  totalTest = 1e6,  # Total possible peaks in genome
  method = "hyperG"  # or "permutation"
)

# For 3+ peak sets
venn_result <- makeVennDiagram(
  Peaks = list(rep1 = peaks1_filtered,
               rep2 = peaks2_filtered,
               rep3 = peaks3_filtered),
  NameOfPeaks = c("Rep1", "Rep2", "Rep3"),
  totalTest = 1e6
)
```

**Key Parameters:**
- `totalTest`: Total number of possible peaks (should be >> largest peak set)
- `method`: "hyperG" (hypergeometric test) or "permutation"
- `connectedPeaks`: How to count connected peaks ("merge", "min", "keepAll")

### Statistical Testing of Peak Overlaps

Use `peakPermTest()` to test whether peak overlaps are statistically
significant, accounting for genomic binding biases:

```{r peak-perm-test, eval = FALSE}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Permutation test for peak overlap significance
perm_result <- peakPermTest(
  rep1 = peaks1_filtered,
  rep2 = peaks2_filtered,
  TxDb = txdb,
  ntimes = 1000,  # Number of permutations
  seed = 123,
  mc.cores = 2
)

# View results
print(perm_result)
```

**Advantages over simple hypergeometric test:**
- Accounts for genomic binding distribution biases
- More appropriate for ChIP-seq data
- Generates random peaks matching observed binding patterns

```{r enrichment-analysis, eval = FALSE}
# Option 1: Using LOLA
library(LOLA)

# Load LOLA database
lola_db <- loadLOLA("path/to/lola_db")

# Run enrichment
enrichment_results <- runLOLA(
  userSets = consensus_peaks,
  universe = background_peaks,
  regionDB = lola_db
)

# Option 2: Using ChIPpeakAnno overlap functions
# Compare with Roadmap Epigenomics data
roadmap_annotated <- annotatePeaksWithRoadmap(
  peaks = consensus_peaks,
  data_type = "enhancer",
  tissue_types = "BRAIN"
)
```

## Step 11: Gene Set Enrichment Analysis

Extract associated genes and perform gene set enrichment using MSigDB, EnrichR,
or GO/KEGG databases.

```{r gene-set-enrichment, eval = FALSE}
# Extract gene names from annotated peaks
annotated_genes <- unique(mcols(annotated_peaks)$gene_name)
annotated_genes <- annotated_genes[!is.na(annotated_genes)]

# Convert to Entrez IDs
library(org.Hs.eg.db)
gene_ids <- mapIds(org.Hs.eg.db, 
                   keys = annotated_genes,
                   column = "ENTREZID",
                   keytype = "SYMBOL")

# GO enrichment
go_enrichment <- getEnrichedGO(
  annotated_peaks,
  orgAnn = "org.Hs.eg.db",
  feature_id_type = "ensembl_gene_id"
)

# Pathway enrichment
pathway_enrichment <- getEnrichedPATH(
  annotated_peaks,
  orgAnn = "org.Hs.eg.db",
  feature_id_type = "ensembl_gene_id"
)

# Using test_enrichment for MSigDB
library(EnrichmentBrowser)
msigdb_enrichment <- test_enrichment(
  gene_set = annotated_genes,
  database = "MSigDB"
)
```

```{r gene-set-enrichment-example}
# Extract genes from annotated peaks
if (length(tf_annotated) > 0 && "gene_name" %in% colnames(mcols(tf_annotated))) {
  annotated_genes <- unique(mcols(tf_annotated)$gene_name)
  annotated_genes <- annotated_genes[!is.na(annotated_genes)]
  
  cat("Genes associated with peaks: ", length(annotated_genes), "\n")
  if (length(annotated_genes) > 0) {
    cat("  Examples: ", paste(head(annotated_genes, 5), collapse = ", "), "\n")
  }
```

## Step 12: Motif Analysis

**Motif analysis** identifies the enriched DNA patterns within peaks to
understand the binding preference of the target protein. This analysis can
reveal sequence motifs that are recognized by transcription factors or
other DNA-binding proteins, providing insights into the molecular mechanisms
of binding.

Extract peak sequences and perform motif discovery/enrichment using
sophisticated tools.

# Phase 6: Visualization and Signal Analysis

> **Overview**: Visualize peak distributions and signal patterns to validate
> annotation results and understand binding patterns. These visualizations
> help identify expected patterns (e.g., promoter enrichment) and detect
> potential issues.

## Step 13: Genomic Distribution Visualization

### Genomic Element Distribution

Use `assignChromosomeRegion()` to summarize peak distribution across genomic
elements (promoters, UTRs, exons, introns, intergenic):

```{r genomic-distribution, eval = FALSE}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Get genomic element distribution
distribution <- assignChromosomeRegion(
  peaks = consensus_peaks,
  nucleotideLevel = FALSE,
  TxDb = txdb,
  tssRegion = c(-3000, 3000)
)

# View distribution percentages
print(distribution$percentage)

# Create pie chart
pie1(distribution$percentage, 
     labels = names(distribution$percentage),
     main = "Peak Distribution Across Genomic Elements")
```

### Metagene Plots

Visualize peak distribution patterns around genomic features using
`metagenePlot()`:

```{r metagene-plot, eval = FALSE}
# Create metagene plot around TSS
metagenePlot(
  peaks = consensus_peaks,
  AnnotationData = txdb,
  PeakLocForDistance = "middle",
  FeatureLocForDistance = "TSS",
  upstream = 3000,
  downstream = 3000
)
```

## Step 14: Signal Analysis and Heatmaps

### Extracting Signals Around Features

Use `featureAlignedSignal()` to extract read coverage or signal around
genomic features for visualization:

```{r feature-aligned-signal, eval = FALSE}
# Load BigWig files or create coverage from BAM files
library(rtracklayer)
bw_file <- "sample.bigWig"
cov <- import(bw_file, format = "BigWig")

# Extract signals around TSS
tss <- promoters(genes(txdb), upstream = 2000, downstream = 2000)
signals <- featureAlignedSignal(
  cvglists = list(sample = cov),
  feature.gr = tss,
  upstream = 2000,
  downstream = 2000,
  n.tile = 100
)

# Create heatmap
featureAlignedHeatmap(signals$sample, 
                     feature.gr = tss,
                     upstream = 2000,
                     downstream = 2000)

# Create average signal profile
featureAlignedDistribution(signals$sample,
                          feature.gr = tss,
                          upstream = 2000,
                          downstream = 2000)
```

**Key Functions:**
- `featureAlignedSignal()`: Extract signals into matrices
- `featureAlignedHeatmap()`: Create heatmaps of signals
- `featureAlignedDistribution()`: Plot average signal profiles

These visualizations help identify binding patterns, such as promoter
enrichment for transcription factors or gene body enrichment for histone marks.

### Sequence Extraction

First, extract sequences from peaks using `getAllPeakSequence()`.

```{r sequence-extraction, eval = FALSE}
# Load genome
library(BSgenome.Hsapiens.UCSC.hg38)
genome <- BSgenome.Hsapiens.UCSC.hg38

# Extract sequences (extend peaks by 50bp on each side for motif analysis)
peak_sequences <- getAllPeakSequence(
  consensus_peaks,
  upstream = 50,
  downstream = 50,
  genome = genome
)

# Write to FASTA for external tools
write2FASTA(peak_sequences, 
            file = "peak_sequences.fasta")
```

### Motif Analysis with R Packages

#### Using memes Package

```{r motif-memes, eval = FALSE}
library(memes)

# Run STREME for de novo motif discovery
streme_results <- runStreme(
  input = peak_sequences,
  control = background_sequences,  # Random genomic regions
  outdir = "streme_results"
)

# Run FIMO for known motif scanning
library(JASPAR2020)
pfm_list <- getMatrixSet(JASPAR2020, opts = list(species = "Homo sapiens"))

fimo_results <- runFimo(
  peaks = consensus_peaks,
  motifs = pfm_list,
  genome = genome
)

# Visualize motifs
plotMotifStack(streme_results$motifs)
```

#### Using monaLisa Package

The `monaLisa` package provides sophisticated motif enrichment analysis with
sequence composition correction, similar to HOMER. It offers several
approaches for identifying enriched motifs:

**1. Binned Motif Enrichment Analysis** (Recommended)

For regions with associated numerical values (e.g., fold change, signal
intensity), group regions into bins and calculate motif enrichments for each
bin:

```{r motif-monalisa-binned, eval = FALSE}
library(monaLisa)
library(JASPAR2020)
library(Biostrings)

# Load JASPAR motifs
pfm_list <- getMatrixSet(JASPAR2020, 
                         opts = list(species = "Homo sapiens",
                                     collection = "CORE"))

# Convert sequences to DNAStringSet
peak_seqs <- DNAStringSet(peak_sequences$sequence)
names(peak_seqs) <- paste0("peak_", seq_along(peak_seqs))

# Example: Bin peaks by fold change (if available)
# peak_fc <- mcols(consensus_peaks)$foldChange
# peak_bins <- bin(x = peak_fc, binmode = "equalN", nElement = 400)

# For demonstration: simple two-bin comparison
# (peaks vs all other peaks as background)
peak_bins <- factor(rep("peaks", length(peak_seqs)))

# Calculate motif enrichment with sequence composition correction
enrichment_se <- calcBinnedMotifEnrR(
  seqs = peak_seqs,
  bins = peak_bins,
  pwmL = pfm_list,
  background = "genome",  # Use genomic background
  genome = BSgenome.Hsapiens.UCSC.hg38,
  BPPARAM = BiocParallel::SerialParam()
)

# The result is a SummarizedExperiment with multiple assays:
# - negLog10P: raw P-values
# - negLog10Padj: adjusted P-values (FDR)
# - pearsonResid: standardized Pearson residuals
# - log2enr: log2 enrichment
# - expForegroundWgtWithHits: expected foreground weight with hits

# Visualize results
plotMotifHeatmaps(enrichment_se, which.plots = "enr")
```

**2. Two-Set Comparison** (Foreground vs Background)

Compare motif enrichments between two sets of sequences:

```{r motif-monalisa-twoset, eval = FALSE}
# Define foreground (peaks of interest) and background sets
foreground_seqs <- peak_seqs[1:500]  # Example: subset of peaks
background_seqs <- peak_seqs[501:1000]  # Example: other peaks

# Create bins for two-set comparison
seq_bins <- factor(c(rep("foreground", length(foreground_seqs)),
                      rep("background", length(background_seqs))))

# Combine sequences
all_seqs <- c(foreground_seqs, background_seqs)

# Calculate enrichment
enrichment_se <- calcBinnedMotifEnrR(
  seqs = all_seqs,
  bins = seq_bins,
  pwmL = pfm_list,
  background = "otherBins",  # Use other bins as background
  BPPARAM = BiocParallel::SerialParam()
)

# Visualize
plotMotifHeatmaps(enrichment_se, which.plots = "enr")
```

**3. Single Set vs Genomic Background**

Compare a single set of sequences against a matched genomic background:

```{r motif-monalisa-single, eval = FALSE}
# Calculate enrichment against genomic background
enrichment_se <- calcBinnedMotifEnrR(
  seqs = peak_seqs,
  bins = factor(rep("peaks", length(peak_seqs))),
  pwmL = pfm_list,
  background = "genome",  # Use genomic background
  genome = BSgenome.Hsapiens.UCSC.hg38,
  BPPARAM = BiocParallel::SerialParam()
)

# Extract significantly enriched motifs
sig_motifs <- enrichment_se[
  rowData(enrichment_se)$negLog10Padj[, "peaks"] > 2,  # FDR < 0.01
]

# Visualize top enriched motifs
plotMotifHeatmaps(enrichment_se[1:20, ], which.plots = "enr")
```

**Key Features of monaLisa:**

- **Sequence composition correction**: Similar to HOMER, corrects for GC content
  and sequence composition biases
- **Multiple statistical measures**: Provides P-values, adjusted P-values,
  Pearson residuals, and enrichment scores
- **Flexible background options**: Can use genomic background, other bins, or
  custom background sequences
- **Efficient implementation**: Faster than HOMER for large datasets
- **Integration with Bioconductor**: Works seamlessly with GRanges, DNAStringSet,
  and other Bioconductor classes

For more details, see the [monaLisa
vignette](https://bioconductor.org/packages/release/bioc/vignettes/monaLisa/inst/doc/monaLisa.html).
```


### Motif Analysis with External Tools

#### Using HOMER

```{r motif-homer, eval = FALSE}
# Write sequences to FASTA
write2FASTA(peak_sequences, file = "peaks.fasta")

# HOMER commands (run in terminal):
# findMotifsGenome.pl peaks.bed hg38 homer_output/ -size 200
# findMotifs.pl peaks.fasta fasta homer_output/ -fastaBg background.fasta

# Parse HOMER results in R
homer_motifs <- read.table("homer_output/knownResults.txt", 
                          sep = "\t", 
                          header = TRUE)
```

#### Using STREME (MEME Suite)

```{r motif-streme, eval = FALSE}
# Write sequences to FASTA
write2FASTA(peak_sequences, file = "peaks.fasta")
write2FASTA(background_sequences, file = "background.fasta")

# STREME command (run in terminal):
# streme --p peaks.fasta --n background.fasta --oc streme_output

# Parse STREME results
# STREME outputs TSV files that can be read into R
streme_motifs <- read.table("streme_output/streme.txt",
                           sep = "\t",
                           header = TRUE)
```

#### Using FIMO (MEME Suite)

```{r motif-fimo, eval = FALSE}
# Prepare motif database (JASPAR, TRANSFAC, etc.)
# Download motifs in MEME format

# FIMO command (run in terminal):
# fimo --oc fimo_output --thresh 1e-4 motifs.meme peaks.fasta

# Parse FIMO results
fimo_results <- read.table("fimo_output/fimo.tsv",
                          sep = "\t",
                          header = TRUE,
                          comment.char = "")

# Filter significant matches
sig_matches <- fimo_results[fimo_results$p.value < 1e-4, ]
```

### Integrated Motif Analysis Workflow

```{r motif-integrated, eval = FALSE}
# Complete motif analysis workflow
library(memes)
library(monaLisa)
library(JASPAR2020)

# 1. Extract sequences
peak_sequences <- getAllPeakSequence(
  consensus_peaks,
  upstream = 100,
  downstream = 100,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

# 2. De novo discovery with STREME (via memes)
streme_results <- runStreme(
  input = peak_sequences,
  control = getRandomSequences(genome, n = length(peak_sequences)),
  outdir = "streme_output"
)

# 3. Known motif enrichment with monaLisa
library(monaLisa)
library(JASPAR2020)
library(Biostrings)

pfm_list <- getMatrixSet(JASPAR2020, 
                         opts = list(species = "Homo sapiens",
                                     collection = "CORE"))

# Convert sequences to DNAStringSet
peak_seqs <- DNAStringSet(peak_sequences$sequence)
names(peak_seqs) <- paste0("peak_", seq_along(peak_seqs))

# Calculate enrichment with sequence composition correction
enrichment_se <- calcBinnedMotifEnrR(
  seqs = peak_seqs,
  bins = factor(rep("peaks", length(peak_seqs))),
  pwmL = pfm_list,
  background = "genome",  # Use genomic background
  genome = BSgenome.Hsapiens.UCSC.hg38,
  BPPARAM = BiocParallel::SerialParam()
)

# 4. Visualize results
plotMotifHeatmaps(enrichment)
plotMotifStack(streme_results$motifs)

# 5. Integrate with peak annotations
# Link motifs to genes through peak annotations
motif_gene_associations <- merge(
  fimo_results,
  mcols(annotated_peaks),
  by.x = "sequence_name",
  by.y = "peak_name"
)
```

# Complete Workflow Example

Here's a complete example workflow for a transcription factor ChIP-seq
experiment:

```{r complete-workflow, eval = FALSE}
# ============================================
# COMPLETE WORKFLOW EXAMPLE
# ============================================

# Setup
library(ChIPpeakAnno)
library(EnsDb.Hsapiens.v86)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(BSgenome.Hsapiens.UCSC.hg38)

# Set global options
setChIPpeakAnnoGlobals(
  genome = "hg38",
  species = "Homo sapiens",
  TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
  EnsDb = EnsDb.Hsapiens.v86
)

# ============================================
# PHASE 1: Per-Replicate Processing
# ============================================

# Step 1: Import peaks
rep1 <- toGRanges("rep1_peaks.bed", format = "BED")
rep2 <- toGRanges("rep2_peaks.bed", format = "BED")
rep3 <- toGRanges("rep3_peaks.bed", format = "BED")

# Step 2: Quality assessment
rep1_assessment <- assessPeaks(rep1, plot = TRUE)
rep2_assessment <- assessPeaks(rep2, plot = TRUE)
rep3_assessment <- assessPeaks(rep3, plot = TRUE)

# Step 3: Filter peaks
rep1_filtered <- filterPeaks(rep1, minWidth = 100, maxWidth = 2000)
rep2_filtered <- filterPeaks(rep2, minWidth = 100, maxWidth = 2000)
rep3_filtered <- filterPeaks(rep3, minWidth = 100, maxWidth = 2000)

# ============================================
# PHASE 2: Consensus Peak Creation
# ============================================

peak_list <- GRangesList(
  rep1 = rep1_filtered,
  rep2 = rep2_filtered,
  rep3 = rep3_filtered
)

consensus_result <- findOverlapsOfPeaks(peak_list, minoverlap = 2)
consensus_peaks <- consensus_result$mergedPeaks

# ============================================
# PHASE 3: Annotation
# ============================================

# Step 5: Preliminary annotation
prelim <- getGenomicAnnotation(consensus_peaks, tssRegion = c(-3000, 3000))

# Step 6: Hierarchical annotation
annotated_peaks <- annotateHierarchically(
  peaks = consensus_peaks,
  factor_type = "TF",
  PeakLocForDistance = "middle",
  bindingRegion = c(-3000, 3000)
)

# ============================================
# PHASE 4: Enhancer Identification
# ============================================

# Annotate with cCRE data
enhancer_annotated <- annotatePeaksWithcCRE(
  peaks = annotated_peaks,
  tissue_types = "any"
)

# ============================================
# PHASE 5: Downstream Analysis
# ============================================

# Step 10: Enrichment analysis
roadmap_annotated <- annotatePeaksWithRoadmap(
  peaks = consensus_peaks,
  data_type = "enhancer",
  tissue_types = "BRAIN"
)

# Step 11: Gene set enrichment
annotated_genes <- unique(mcols(annotated_peaks)$gene_name)
annotated_genes <- annotated_genes[!is.na(annotated_genes)]

library(org.Hs.eg.db)
go_enrichment <- getEnrichedGO(
  annotated_peaks,
  orgAnn = "org.Hs.eg.db"
)

# Step 12: Motif analysis
library(memes)
library(monaLisa)

# Extract sequences
peak_sequences <- getAllPeakSequence(
  consensus_peaks,
  upstream = 100,
  downstream = 100,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

# De novo discovery with STREME (via memes)
streme_results <- memes::runStreme(
  input = peak_sequences,
  control = getRandomSequences(BSgenome.Hsapiens.UCSC.hg38, 
                               n = length(peak_sequences)),
  outdir = "streme_output"
)

# Known motif enrichment with monaLisa
library(JASPAR2020)
library(Biostrings)

pfm_list <- getMatrixSet(JASPAR2020, 
                         opts = list(species = "Homo sapiens",
                                     collection = "CORE"))

# Convert sequences to DNAStringSet for monaLisa
peak_seqs <- DNAStringSet(peak_sequences$sequence)
names(peak_seqs) <- paste0("peak_", seq_along(peak_seqs))

# Calculate enrichment (single set vs genomic background)
enrichment_se <- monaLisa::calcBinnedMotifEnrR(
  seqs = peak_seqs,
  bins = factor(rep("peaks", length(peak_seqs))),
  pwmL = pfm_list,
  background = "genome",
  genome = BSgenome.Hsapiens.UCSC.hg38,
  BPPARAM = BiocParallel::SerialParam()
)

# Visualize
plotMotifHeatmaps(enrichment)
```

# Best Practices and Common Pitfalls

This section summarizes key best practices for peak annotation. For
comprehensive guidelines, see the [Best Practices
Document](docs/Peak_Annotation_Best_Practices.md), which covers community
standards, quality control, annotation strategies, and factor-specific
recommendations.

## Best Practices

### Pre-Annotation Considerations

1. **Always assess quality first**: Use `assessPeaks()` on each replicate before
   proceeding. Check FRiP scores, peak widths, and score distributions.

2. **Match annotation to genome assembly**: Always ensure your annotation data
   matches your peak genome assembly. Use `seqlevelsStyle()` to align
   chromosome naming if needed.

3. **Use latest annotation versions**: Prefer annotations (GENCODE, Ensembl) and
   document annotation version numbers for reproducibility.

### Annotation Strategy Selection

4. **Match strategy to factor type**: Use appropriate `factor_type` in
   `annotateHierarchically()` or select appropriate parameters for
   `annotatePeakInBatch()` based on your binding factor:
   - **TFs**: Promoter-focused (`bindingRegion = c(-2000, 500)`)
   - **H3K4me3**: TSS-focused (`bindingRegion = c(-2000, 500)`)
   - **H3K27ac**: Promoter + Enhancer (`bindingRegion = c(-5000, 3000)`)
   - **H3K36me3**: Gene body (`bindingType = "fullRange"`)
   - **H3K27me3**: Broad domains (`bindingType = "fullRange"`)

5. **Use appropriate reference points**: 
   - For narrow peaks (<200bp): `PeakLocForDistance = "start"` (default)
   - For broad peaks (>200bp): `PeakLocForDistance = "middle"` (recommended)
   - For strand-specific assays: `PeakLocForDistance = "endMinusStart"`

6. **Follow ENCODE standards for distance cutoffs**:
   - Proximal promoter: -2000 to +500 bp from TSS
   - Core promoter: -500 to +100 bp from TSS
   - Extended promoter: -5000 to +3000 bp from TSS
   - Enhancer: >5 kb from TSS

### Quality Control

7. **Validate annotation results**: Always validate against known biological
   knowledge. Check that promoter-binding factors show enrichment near TSS,
   gene body factors show enrichment across gene bodies, etc.

8. **Use multiple annotation approaches**: When possible, use multiple
   annotation methods and compare results to increase confidence.

9. **Filter appropriately**: Don't over-filter; use assessment results to guide
   filtering. Maintain biological signal while removing low-quality peaks.

### Workflow Best Practices

10. **Use global options**: Set genome, species, TxDb, and EnsDb globally for
    convenience using `setChIPpeakAnnoGlobals()`.

11. **Consider replicates**: Always use multiple replicates and create consensus
    peaks if possible. Use IDR for 2 replicates, majority voting for 3+.

12. **Document your workflow**: Record all parameters, annotation versions, and
    genome assemblies used for reproducibility.

13. **For DNA-seq methods**: Always use `ignore.strand = TRUE` (default).

14. **For RNA-seq methods**: Use `ignore.strand = FALSE` only if strandedness
    information is preserved by the sequencing protocol and maintained in the
    peak data.

## Common Pitfalls

1. **Wrong reference point**: Using `PeakLocForDistance="start"` for broad peaks
   (>200bp) can give misleading distances. Use `"middle"` for peaks >200bp.

2. **Ignoring factor-specific patterns**: Different factors require different
   annotation strategies. Don't use the same parameters for TFs and histone
   marks.

3. **Not assessing quality**: Always assess peak quality before annotating.
   Poor-quality peaks lead to poor annotations.

4. **Genome assembly mismatch**: Using a different genome assembly version for
   annotation than for peaks causes incorrect annotations. Always verify
   assembly matching.

5. **Over-filtering**: Too strict filtering loses biological signal. Use
   assessment results to guide appropriate filtering thresholds.

6. **Single replicate analysis**: Always compare replicates and use IDR or
   majority voting to create consensus peaks.

7. **Ignoring strand information**: For strand-specific assays (e.g., GRO-seq,
   CLIP-seq), ensure proper strand handling with `ignore.strand = FALSE` and
   appropriate `PeakLocForDistance` settings.

8. **Using default parameters without consideration**: Default parameters may not
   be optimal for your data. Consult the [Parameter
   Guide](docs/Peak_Annotation_Parameter_Guide.md) to select appropriate
   parameters.

9. **Not validating results**: Always check annotation distributions and
   validate against known biological patterns (e.g., promoter enrichment for
   TFs).

10. **Inconsistent annotation across conditions**: When comparing conditions,
    use the same annotation parameters and annotation data version for all
    samples.

# Frequently Asked Questions (FAQ)

## Can I output either the nearest or overlapping features?

This question pertains to this
[post](https://support.bioconductor.org/p/9154859/#9154869).

When you set `output = "both"`, both the "nearest" and "overlapping" features
will be output. If your preference is to assign only one type of feature to
each peak (either "nearest" or "overlapping", with a preference for
"overlapping" in cases where the "nearest" feature is not "overlapping"), you
can use the following strategy: first, annotate peaks with overlapping
features; second, annotate peaks lacking overlapping features with the nearest
features; last, concatenate the two sets of results. The example codes are
provided below:

```{r, warning = FALSE}
library(ensembldb)
library(EnsDb.Hsapiens.v75)
data(myPeakList)
annoData <- annoGR(EnsDb.Hsapiens.v75)

# Step1: annotate peaks to the overlapping features, if "select = 'all'", 
# multiple features can be assigned to a single peak.
anno_overlapping <- annotatePeakInBatch(myPeakList, 
                                        AnnotationData = annoData, 
                                        output = "overlapping", 
                                        select = "first")
anno_overlapping_non_na <- anno_overlapping[
  !is.na(anno_overlapping$feature)]

# Step2: annotate peaks that are without overlapping features to nearest 
# features
myPeakList_non_overlapping <- myPeakList[
  !(names(myPeakList) %in% anno_overlapping_non_na$peak)]  
anno_nearest <- annotatePeakInBatch(myPeakList_non_overlapping,
                                    AnnotationData = annoData, 
                                    output = "nearestLocation", 
                                    select = "first")

# Step3: concatenate the two
anno_final <- c(anno_overlapping_non_na, anno_nearest)
```

The provided code allocates either the "overlapping" or "nearest" feature to
each peak. If the "overlapping" feature is not the "nearest", only the
"overlapping" one will be reported.

## Why is the sum of peak numbers in the Venn diagram NOT equal to the sum of the peaks in the original peak lists?

This question is a common scenario in calculating the intersection of peaks.
Peaks, being a range of continuous points rather than single points, present a
challenge in determining the overlapping regions. Consider the intersection of
two lists of range objects: list A (1~3, 4~5, 7~9) and list B (2~8), the
number of overlapping ranges depends on the reference chosen, if we use list B
as the reference, the output would be one range; however, if we use list A as
the reference, the output would be three ranges.

## How does `findOverlapsOfPeaks` count the number of overlapping peaks?

This question is in response to this
[post](https://support.bioconductor.org/p/133486/#133603).

When counting the number of overlapping peaks using `findOverlapsOfPeaks`, the
`connectedPeaks` option comes into play. If multiple peaks involve in any
group of connected or overlapping peaks in any input peak list, setting
`connectedPeaks = "merge"` will increment the overlapping counts by one. On the
other hand, setting `connectedPeaks = "min"` will add the minimal number of
involved peaks in each group of connected or overlapped peaks to the
overlapping counts. If `connectedPeaks = "keepAll"`, it will add the number of
involved peaks for each peak list to the corresponding overlapping counts. In
addition, it will output counts as if `connectedPeaks` was set to "min".

For examples, if 5 peaks in group1 overlap with 2 peaks in group 2:

+ Setting `connectedPeaks = "merge"` will add 1 to the overlapping counts
+ Setting `connectedPeaks = "min"` will add 2 to the overlapping counts
+ Setting `connectedPeaks = "keepAll"` will add 5 peaks to "count.group1", 2 to
  "count.group2", and 2 to the overlapping counts

## Is there a way to show the number of peaks in original peak lists?

You have the option to configure `connectedPeaks = "keepAll"` in both the
`findOverlapsOfPeaks` and `makeVennDiagram` functions.

## How to extract the original peak IDs of the overlapping peaks?

In the output of `findOverlapsOfPeaks`, each element in the peak list contains
a metadata column names _peakNames_, which is a CharacterList. This
CharacterList is a list of the contributing peak IDs with prefixes, e.g.
"peaks1_peakname1", "peaksi_peaknamej", where "i" denotes the peak group i and
"j" denotes the specific peak j within that group. Users can retrieve the
original peak name by splitting these strings. Here are the sample codes:

```{r, warning = FALSE}
library(ChIPpeakAnno)
library(reshape2)

# Step1: read in two peak files
bed <- system.file("extdata", 
                   "MACS_output.bed",
                   package = "ChIPpeakAnno")
gff <- system.file("extdata", 
                   "GFF_peaks.gff", 
                   package = "ChIPpeakAnno")
gr1 <- toGRanges(bed, format = "BED", 
                 header = FALSE)
gr2 <- toGRanges(gff, format = "GFF", 
                 header = FALSE, skip = 3)
names(gr2) <- seq(length(gr2)) # add names to gr2 for Step4

# Step2: find overlapping peaks
ol <- findOverlapsOfPeaks(gr1, gr2)
peakNames <- ol$peaklist[['gr1///gr2']]$peakNames

# Step3: extract original peak names
peakNames <- melt(peakNames, 
                  value.name = "merged_peak_id") # reshape df
head(peakNames, n = 2)

peakNames <- cbind(peakNames[, 1], 
                   do.call(rbind, 
                           strsplit(as.character(peakNames[, 3]), "__")))

colnames(peakNames) <- c("merged_peak_id", "group", "peakName")
head(peakNames, n = 2)

# Step4: split by peak group
gr1_subset <- gr1[peakNames[peakNames[, "group"] %in% "gr1", "peakName"]]
gr2_subset <- gr2[peakNames[peakNames[, "group"] %in% "gr2", "peakName"]]
head(gr1_subset, n = 2)
head(gr2_subset, n = 2)
```

Alternatively, the following snippet should also work.

```{r}
gr1_renamed <- ol$all.peaks$gr1
gr2_renamed <- ol$all.peaks$gr2
head(gr1_renamed, n = 2)
head(gr2_renamed, n = 2)

peakNames <- melt(ol$peaklist[['gr1///gr2']]$peakNames, 
                  value.name = "merged_peak_id")
gr1_subset <- gr1_renamed[peakNames[grepl("^gr1", peakNames[, 3]), 3]]
gr2_subset <- gr2_renamed[peakNames[grepl("^gr2", peakNames[, 3]), 3]]
head(gr1_subset, n = 2)
head(gr2_subset, n = 2)
```

## How to select the proper number for `totalTest` when using `makeVennDiagram`?

During the evaluation of the association between two sets of peak data using
the hypergeometric distribution, it is essential to determine the total number
of potential binding sites, the `totalTest` parameter in the
`makeVennDiagram`. This parameter should have a value larger than the largest
number of peaks in the peak list. The choice of `totalTest` influences the
stringency of the test, with smaller values yielding more stringent tests and
larger P-values. The computation time for calculating P-values remains
independent of the `totalTest` value. For practical guidance on selecting an
appropriate `totalTest` value, please refer to this
[post](https://stat.ethz.ch/pipermail/bioconductor/2010-November/036540.html).

# Session info

Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:

```{r sessionInfo, echo = FALSE}
sessionInfo()
```

# References

The references are automatically generated from the bibliography file
`bibliography.bib` using the citations in the text.

